%% Don't forget to change the paper format in the next line
%
% translated to Japanese by Shigeharu TAKENO
% Original: Id: faq.tex updated at 2019-05-03
%
%%% \documentclass[a4paper,11pt]{article}
\documentclass[a4paper,11pt]{jarticle}
%% end Japanese
\usepackage[margin=2cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[hyphens]{url}
\urlstyle{sf}

\parindent=0pt
\parskip=5pt

\ifx\pdfoutput\undefined
    % latex or latex2html output
    \usepackage{times,mathptmx}
    \usepackage[
        hypertex,
        hyperindex,
        bookmarks,
        bookmarksnumbered=true,
        pdftitle={gnuplot faq},
        pdfauthor={gnuplot},
        pdfsubject={see www.gnuplot.info}
        % , pdfcreator={}
        % , pdfkeywords={...}
    ]{hyperref}
    \newcommand{\latex}[1]{#1}
\else % *** pdflatex output
    \usepackage{times,mathptmx}
    \usepackage[
%       pdftex,
        hyperindex,
        bookmarks,
        bookmarksnumbered=true,
        pdftitle={gnuplot faq},
        pdfauthor={gnuplot},
        pdfsubject={see www.gnuplot.info}
        % , pdfcreator={}
        % , pdfkeywords={...}
    ]{hyperref}
\fi

\usepackage{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\hypersetup{
  colorlinks   = true, %Colours links instead of ugly boxes
  linkcolor    = darkblue, %Colour of internal links
  urlcolor     = blue %Colour for external hyperlinks
}

% There may be incompatibilities between different versions of
% url.sty, html.sty and hyperref.sty -- it seems there are machines which
% cannot combine them together with simultaneous output to dvi, pdf, html.
% Thus do it this way:
\ifx\html\undefined
    % Modified Dec 2014 (EAM) for use with htlatex or pdflatex
    \def\http#1{{\small\href{http://#1}{\url{http://#1}}}}
    \def\mailto#1{{\small\href{mailto://#1}{\url{mailto://#1}}}}
    \def\news#1{\href{news://#1}{\url{news://#1}}}
%%%  \def\ftp#1#2{\href{ftp://#1#2}{\url{#1} in \url{#2}}}
    \def\ftp#1#2{\href{ftp://#1#2}{\url{#1} の \url{#2}}}
%% end Japanese
\else
    % Running this file by latex2html:
    \usepackage{html}
%    \html{
        \newcommand{\news}[1]%
            {\def~{\~{}}\htmladdnormallink{\latex{\url{#1}}\html{\textit{#1}}}%
                {news:#1}%
            }
        \newcommand{\ftp}[2]%
%%%          {\htmladdnormallink{\latex{\url{#1}{} in \url{#2}}%
%%%                  \html{\textit{#1} in \textit{#2}}}%
            {\htmladdnormallink{\latex{\url{#1}{} の \url{#2}}%
                    \html{\textit{#1} の \textit{#2}}}%
%% end Japanese
                {ftp://#1#2}%
            }
        \newcommand{\mailto}[1]%
            {\htmladdnormallink{\latex{\url{<#1>}}\html{\textit{#1}}}%
                {mailto:#1}%
            }
        \newcommand{\http}[1]%
            {\htmladdnormallink{\latex{\url{http://#1}}%
                    \html{\textit{http://#1}}}%
                {http://#1}%
            }
%   }
\fi


% History:
% version 1.4 dated 99/10/07
% Revision 1.58 dated 2017/06/04 (last CVS version)
%

%
\newcommand{\gnuplot}{\textbf{gnuplot }}
\newcommand{\Gnuplot}{\textbf{Gnuplot }}

\begin{document}

\title{\gnuplot FAQ}
\author{}
\date{}
\maketitle

\noindent
%%% This material describes \gnuplot version 5
この文書は、\gnuplot バージョン 5 についての記述です。
%% end Japanese
        \\
%%% FAQ version: April 2019
FAQ バージョン: 2019 4 月


\tableofcontents

\newpage
\setcounter{section}{-1}
%%% \section{Meta -- Questions}
\section{メタな質問}

%%% \subsection{Where do I get this document?}
\subsection{このドキュメントはどこで手に入れられますか}

%%% The newest version of this document is on the web at
%%% \http{www.gnuplot.info/faq/}.
このドキュメントの最新版は、Web 上の %
\http{www.gnuplot.info/faq/} %
にあります。

%%% \subsection{Where do I send comments about this document?}
\subsection{このドキュメントに関するコメントはどこに送れば良いですか}

%%% Send comments, suggestions etc to the developer mailing list
%%% \mailto{gnuplot-beta@lists.sourceforge.net}.
コメント、提案等は開発者向けメーリングリスト %
\mailto{gnuplot-beta@lists.sourceforge.net} に送ってください。

%%% \section{General Information}
\section{一般的な質問}

%%% \subsection{What is \gnuplot?}
\subsection{\gnuplot とは何ですか}

%%% \gnuplot is a command-driven plotting program.
%%% It can be used interactively to plot functions and data points in
%%% both two- and three-dimensional plots in many different styles and
%%% many different output formats.  \Gnuplot can also be used as a
%%% scripting language to automate generation of plots.
%%% It is designed primarily for the visual display of scientific data.
\gnuplot はコマンド駆動型の描画プログラムです。関数の描画、
およびデータ点の描画を、2 次元、3 次元の両方で、
そして多くの異なるスタイル、多くの異なる出力形式で行うことができます。
また、\gnuplot は自動的にグラフを生成するスクリプト言語としても
使用することができます。
これは主に科学的なデータを画面表示するものとして設計されています。
%% end Japanese
%%% \gnuplot is copyrighted, but freely distributable;
%%% you don't have to pay for it. You are welcome to download the source code.
\gnuplot の著作権は保護されていますが、自由に配布でき、
代金を払う必要はありません。ソースコードのダウンロードも歓迎します。


%%% \subsection{How did it come about and why is it called \gnuplot?}
\subsection{それはどこから来て、そしてなぜ \gnuplot と呼ばれているのですか}

%%% The authors of \gnuplot are:
%%% Thomas Williams, Colin Kelley, Russell Lang, Dave Kotz, John
%%% Campbell, Gershon Elber, Alexander Woo and many others.
\gnuplot の作者は Thomas Williams, Colin Kelley, Russell Lang,
Dave Kotz, John Campbell, Gershon Elber, Alexander Woo、
そして多くの協力者です。

%%% The following quote comes from Thomas Williams:
以下は Thomas Williams によります:
%% end Japanese
\begin{quote}
%%%      I was taking a differential equation class and Colin was taking
%%%      Electromagnetics, we both thought it'd be helpful to visualize the
%%%      mathematics behind them. We were both working as sys admin for an
%%%      EE VLSI lab, so we had the graphics terminals and the time to do
%%%      some coding. The posting was better received than we expected, and
%%%      prompted us to add some, albeit lame, support for file data.
     私は微分方程式の講義を、そして Colin は電磁気学の講義を取っていて、
     2 人ともそれらに関する数学が視覚化できればなあと考えていました。
     私達はある EE (Engineering Electronics) VLSI 研の
     システム管理者として働いていて、そして、
     グラフィック端末とコーディングを行なう時間がありました。
     その投稿は我々の期待以上に良く受け入れられ、そして、不完全でしたが、
     ファイルデータに関するいくつかのサポートに我々を駆り立てました。

%%%      Any reference to GNUplot is incorrect. The real name of the program
%%%      is "gnuplot". You see people use "Gnuplot" quite a bit because many
%%%      of us have an aversion to starting a sentence with a lower case
%%%      letter, even in the case of proper nouns and titles. gnuplot is not
%%%      related to the GNU project or the FSF in any but the most
%%%      peripheral sense. Our software was designed completely
%%%      independently and the name "gnuplot" was actually a compromise. I
%%%      wanted to call it "llamaplot" and Colin wanted to call it "nplot."
%%%      We agreed that "newplot" was acceptable but, we then discovered
%%%      that there was an absolutely ghastly pascal program of that name
%%%      that the Computer Science Dept.\ occasionally used. I decided that
%%%      "gnuplot" would make a nice pun and after a fashion Colin agreed.
     GNUplot という紹介はどんなものも正しくはありません。
     このプログラムの本当の名前は "gnuplot" です。
     ずいぶんと "Gnuplot" と書かれているのを見るでしょうが、
     それは我々の大半が、固有名詞やタイトルであっても、
     文章を小文字で開始することに嫌悪を持っているからです。
     gnuplot は GNU プロジェクトや FSF とは、
     極わずかの意味を除けば少しも関係ありません。
     我々のソフトウェアは完全に独立にデザインされたもので、
     そして "gnuplot" という名前は実は妥協によるものです。
     私は "llamaplot" と呼びたかったし、
     Colin は "nplot" と呼びたかったのです。
     そして、我々は "newplot" がいいだろうということで合意しましたが、
     しかしそのときその名前を持つ、絶対的におぞましい Pascal のプログラムを
     計算機科学部でたまに使っていることを知りました。
     それで私は "gnuplot" がいいごろ合わせになるとして、
     後で Colin もそれに同意したのです。
%% end Japanese
\end{quote}

 
%%% \subsection{What does \gnuplot offer?}
\subsection{\gnuplot は何を提供してくれますか}

\begin{itemize}
%%% \item Two-dimensional functions and data plots combining many different
%%%  elements such as points, lines, error bars, filled shapes, labels, 
%%%  arrows, ...
%%% \item Polar axes, log-scaled axes, general nonlinear axis mapping, 
%%%  parametric coordinates
%%% \item Data representations such as heat maps, beeswarm plots, violin 
%%%  plots, histograms, ...
%%% \item Three-dimensional plots of data points, lines, and surfaces in
%%%  many different styles (contour plot, mesh)
%%% \item Algebraic computation using integer, floating point, or complex
%%%  arithmetic
%%% \item Data-driven model fitting using Marquardt-Levenberg minimization
%%% \item Support for a large number of operating systems, graphics
%%%  file formats and output devices
%%% \item Extensive on-line help
%%% \item \TeX{}-like text formatting for labels, titles, axes, data points
%%% \item Interactive command line editing and history
\item 2 次元での関数やデータの、点、折れ線、誤差表示、塗り潰した図形、
ラベル文字列、ベクトルなどの沢山の異なる要素を組み合わせた描画
\item 極座標軸、対数軸、より一般の非線形軸の割り当てや媒介変数座標
\item 温度分布、ビースウォームグラフ、バイオリングラフ、ヒストグラムなどの
データ表現
\item 3 次元でのデータの、点、折れ線、曲面の、たくさんのスタイル %
(等高線描画、網かけ) による描画
\item 整数、実数、複素数での代数演算
\item Marquardt-Levenberg 最適法を用いたデータ駆動型の曲線回帰
\item 多くのオペレーティングシステム、多くのグラフィックファイル形式出力、
そして多くの出力装置のサポート
\item 広範囲にわたるオンラインヘルプ
\item ラベル、タイトル、軸、データ点への \TeX{} ライクな書式による見出し付け
\item 対話型入力形式のエディット機能とヒストリ (履歴) 機能のサポート
%% end Japanese
\end{itemize}


%%% \subsection{Is \gnuplot suitable for scripting?}
\subsection{\gnuplot はスクリプトに適していますか}

%%% Yes. Gnuplot can read in files containing additional commands during
%%% an interactive session, or it can be run in batch mode by piping a
%%% pre-existing file or a stream of commands to stdin. Gnuplot is used
%%% as a back-end graphics driver by higher-level mathematical packages 
%%% such as Octave and can easily be wrapped in a cgi script for use as a 
%%% web-driven plot generator.  Gnuplot supports context- or data-driven
%%% flow control and iteration using familiar statements
%%% {\em if else continue break while for}.
ええ。gnuplot は対話型実行中に
追加コマンドを含むファイルを読み込むことができますし、
既に存在するファイルや標準入力からのコマンド列を
パイプを使ってバッチモードでそれを処理することもできます。
gnuplot は、Octave のような高水準な数学パッケージの背後で実行する
グラフィックドライバとして使われていますし、
cgi スクリプトでラップすることで
容易に Web 起動型の描画生成ツールとして使うこともできます。
gnuplot は、なじみのある命令 {\em if else continue break while for} を
用いた文脈駆動型、データ駆動型の処理制御、
および繰り返しをサポートしています。


%%% \subsection{Can I run \gnuplot on my computer?}
\subsection{\gnuplot は私のコンピュータで動かすことができますか}

%%% \Gnuplot{} is in widespread use on many platforms, including
%%% MS Windows, linux, unix, and OSX.  The current source code retains
%%% supports for older systems as well, including VMS, Ultrix, OS/2, and
%%% MS-DOS. 16-bit platforms are no longer supported.
\gnuplot は、MS Windows, linux, unix, OSX などの
多くのプラットホーム上で、広く利用できます。
現在のソースコードは、VMS, Ultrix, OS/2, %
MS-DOS などの古いシステムのサポートも保持していますが、
16 ビットプラットホームはもうサポートしていません。

%%% You should be able to compile the \gnuplot source more or
%%% less out of the box in any reasonably standard (ANSI/ISO C, POSIX)
%%% environment.
\gnuplot のソースは、それなりに標準的 (ANSI/ISO C, POSIX 等) な環境から
多少外れたものの上でもコンパイルできるでしょう。


%%% \subsection{Legalities}
\subsection{ライセンスについて}

%%% \Gnuplot{} is authored by a collection of volunteers, who cannot
%%% make any legal statement about the compliance or non-compliance of
%%% \gnuplot or its uses. There is no warranty whatsoever. Use at your
%%% own risk.
\gnuplot はボランティアの集団によって作られたものです。
彼らは \gnuplot、およびその使用の許可、
不許可に関する法的な文書を作成できません。また何の保証もありません。
自分自身の責任で使用してください。

%%% \Gnuplot{} is freeware in the sense that you don't have to pay for it.
%%% You can use or modify gnuplot as you like, however certain restrictions
%%% apply to further distribution of modified versions.
%%% Please read and abide by the modification and redistribution terms in
%%% the \textit{Copyright} file.
\gnuplot は、無償であると言う意味でフリーウェアです。
gnuplot をあなたが望むように使用し、改変することができますが、
改変したものを配布することについてはある種の制限がつきます。
\textit{Copyright} というファイルの modification (改変) と
redistribution (再配布) の項目を読んでそれに従ってください。

%%% \subsection{Does \gnuplot have anything to do with the FSF and the GNU
%%% project?}
\subsection{\gnuplot は FSF や GNU プロジェクトと何か関係があるのですか}

%%% \Gnuplot{} is neither written nor maintained by the FSF. At one
%%% time it
%%% was distributed by the FSF but this is no longer true. \Gnuplot{} 
%%% as a whole
%%% is not covered by the GNU General Public License (GPL).
\gnuplot は FSF が作ったのでも FSF がメンテナンスしているのでもありません。
一時期 FSF から配布されていたこともありますが、
しかし現在はこれは正しくありません。
\gnuplot は、GNU General Public License (GPL) に全く保護されていません。


%%% \subsection{Where do I get further information?}
\subsection{更なる情報はどこで手に入りますか}

%%% See the main gnuplot web page \http{www.gnuplot.info}.
gnuplot Web ページ \http{www.gnuplot.info} を見てください。

%%% Some documentation and tutorials are available in other languages 
%%% than English.
%%% See \http{gnuplot.sourceforge.net/help.html}, section "Localized 
%%% learning pages
%%% about gnuplot", for the most up-to-date list.
英語以外の他の言語での文書や入門書もあります。
最新のリストについては、\http{gnuplot.sourceforge.net/help.html} の %
"Localized learning pages about gnuplot" の節をご覧ください。


%%% \section{Setting it up}
\section{インストール}

%%% \subsection{What is the current version of \gnuplot?}
\subsection{\gnuplot の最新バージョンは}

%%% The current stable version of \gnuplot is 5.2, first released in 
%%% September 2017.
%%% Incremental versions (patchlevel 1, 2, ...) are typically released 
%%% every six months.
%%% The development version of \gnuplot is currently 5.3.
\gnuplot の最新の安定版は、2017 年 9 月に最初にリリースされた 5.2 です。
更新バージョン (パッチレベル 1, 2, ...) は、だいたい 6 ヶ月毎に
リリースされます。
\gnuplot の開発版は、現在は 5.3 となっています。

%%% \subsection{Where can I get \gnuplot?}
\subsection{\gnuplot はどこで手に入れられますか}
%% end Japanese
\label{where-get-gnuplot}

%%% The best place to start is \http{www.gnuplot.info}. From there
%%% you find various pointers to other sites, including the project
%%% development site on SourceForge \http{sourceforge.net/projects/gnuplot}.
初めに最も適切な場所は \http{www.gnuplot.info} です。
ここから、例えば SourceForge 上の開発プロジェクトサイト %
\http{sourceforge.net/projects/gnuplot} などの、
色々なサイトの場所を見つけられるでしょう。

%%% The source distribution ("gnuplot-5.2.6.tar.gz" or a similar name) is
%%% available from the official distribution site 
%%% \http{sourceforge.net/projects/gnuplot}.
ソース配布物 ("gnuplot-5.2.6.tar.gz" または同様の名前) %
は公式の配布サイト \http{sourceforge.net/projects/gnuplot} にあります。

%%% \subsection{Why would I care about the development version?}
\subsection{なぜ開発版を気にしなければいけませんか}

%%% The current development version will generally include features that are
%%% not yet part of the most recent stable release of gnuplot.
%%% As of May 2019 the 5.3 development version notably supports
一般的に、現在の開発版は、最終安定リリース版の gnuplot には
まだ入っていない機能も入っています。
2019 年 4 月現在、開発版のバージョン 5.3 は、以下を特にサポートしています:
%% end Japanese
\begin{itemize}{}{}
\setlength\itemsep{-4pt}
%%% \item 64-bit integer arithmetic
\item 64 ビット整数計算
%% end Japanese
%%% \item 3D voxel grids
\item 3 次元ボクセル格子 (voxel grid)
%% end Japanese
%%% \item 3D plot styles {\em isosurface, circles, boxes}
\item 3 次元描画スタイル {\em isosurface, circles, boxes}
%% end Japanese
%%% \item {\em set walls} command to show xy xz and yz planes in 3D plots
\item コマンド {\em set walls} で 3 次元グラフに xy, xz, yz 平面を表示
%% end Japanese
%%% \item unicode escape sequences {\textbackslash}U+hhhh
\item Unicode エスケープシーケンス {\textbackslash}U+hhhh
%% end Japanese
%%% \item order N Bessel functions of the 1st and 2nd kind
\item 第 1 種、第 2 種の N 次ベッセル関数
%% end Japanese
%%% \item modified Bessel functions
\item 変形ベッセル関数
%% end Japanese
%%% \item multiple textbox styles
\item 複数の textbox スタイル
%% end Japanese
\end{itemize}

%%% \subsection{Where can I get current development version of \gnuplot?}
\subsection{\gnuplot の最新の開発版はどこで手に入れられますか}

%%% The development version of gnuplot is held in a git repository which you
%%% can clone as shown below to inspect or build an executable program
%%% from the source.
gnuplot の開発版は、git リポジトリに保持されていて、以下に示すように
クローンを手元で作成し、ソースを調べたり、
実行プログラムをソースから作成したりできます。

\latex{\scriptsize}
\begin{verbatim}
  git clone https://git.code.sf.net/p/gnuplot/gnuplot-main gnuplot
\end{verbatim}
\latex{\normalsize}

%%% Questions related to the development version should go to
%%% \mailto{gnuplot-beta@lists.sourceforge.net}.
開発版に対する質問は必ず %
\mailto{gnuplot-beta@lists.sourceforge.net} に送ってください。


%%% \subsection{How do I compile \gnuplot on my system?}
\subsection{\gnuplot は私のシステムではどうコンパイルすれば良いですか}

%%% Read the release notes and files \textit{README} and \textit{INSTALL}.
%%% You will need C and C++ compilers and installed versions of various
%%% support libraries depending on what configuration options you choose 
%%% and what terminal types you want your executable to support.
リリースノートと \textit{README}, \textit{INSTALL} のファイルを
読んでください。C と C++ コンパイラが必要で、様々なサポートライブラリが
あらかじめインストールされている必要もありますが、
それはどのような設定オプションを選択するかによりますし、
作成する実行ファイルにどの出力形式をサポートさせたいかにもよります。

\begin{itemize}
\item
%%% To build from a release version on linux, use \textit{./configure}
%%% (or \textit{./configure {-}{-}prefix=\$HOME/usr}
%%% for an installation for a single user), \textit{make} and finally
%%% \textit{make install}.  
linux 上でリリース版からコンパイルするには、\textit{./configure}
(一人の一般ユーザ用にインストールする場合は、
\textit{./configure {-}{-}prefix=\$HOME/usr}) とし、
次に \textit{make} として、最後に \textit{make install} です。
%% end Japanese
%%% Pay close attention to the output from the \textit{configure} script
%%% (yes there is a lot of it).
%%% There you will find hints as to what additional support libraries 
%%% may be needed and what additional options may be desirable.  
%%% In general you will need to have installed the "development" package
%%% for each of the support libraries.
\textit{configure} スクリプトの出力には十分注意を払ってください 
(そう、それはとても長いのだけれど)。
そこには、どのサポートライブラリが必要なのか、
どの追加オプションが望ましいのか、
といったことに対するヒントが現れています。
一般には、各サポートライブラリの「開発版」パッケージを
インストールしておく必要があるでしょう。
%% end Japanese
\item
%%% To build from the development source on linux you must run a separate
%%% script \textit{./prepare}
%%% prior to running \textit{./configure} .
linux 上で開発版ソースからコンパイルするには、
\textit{./configure} を実行させる前に、
別のスクリプト \textit{./prepare} を実行する必要があります。
%% end Japanese
\item
%%% On Windows, makefiles can be found in \textit{config/mingw}, 
%%% \textit{config/msvc},
%%% \textit{config/watcom}, and \textit{config/cygwin}. Update the options 
%%% in the 
%%% makefile's header and run the appropriate \textit{make} tool in the same 
%%% directory
%%% as the makefile. Additional instructions can be found in the makefiles.
Windows では、makefile は \textit{config/mingw}, \textit{config/msvc},
\textit{config/watcom}, \textit{config/cygwin} 内にあります。
makefile の先頭に書かれているオプションを修正し、
その makefile が置かれているのと同じディレクトリ内で
適当な \textit{make} ツールを実行してください。
追加説明も、makefile に書かれています。
%% end Japanese
\item
%%% For other platforms, copy the relevant makefile (e.g. 
%%% \textit{makefile.os2} for 
%%% OS/2) from \textit{config/} to \textit{src/}, optionally update options 
%%% in the 
%%% makefile's header, then change directory to \textit{src} and run 
%%% \textit{make}.
他のプラットフォームでは、適切な makefile 
(例えば OS/2 なら \textit{makefile.os2}) %
を \textit{config/} ディレクトリから \textit{src/} にコピーして、
makefile のヘッダにあるオプションを必要なら更新して、
\textit{src} にディレクトリを移動して、
\textit{make} を実行してください。
%% end Japanese
\end{itemize}


%%% \subsection{What documentation is there, and how do I get it?}
\subsection{ドキュメントはどこですか、それはどうすれば手に入れられますか}

%%% Full documentation is included in the release distribution as a PDF file.
%%% Individual sections can be browsed from inside a gnuplot session
%%% by typing \textit{help {\em keyword}}.
%%% Documentation in other formats can be compiled from source in the
%%% {\em docs} subdirectory.
ソース配布物に完全なドキュメントが PDF ファイル形式で含まれています。
個々のセクションは、gnuplot の作業中に \textit{help {\em キーワード}} を
打つことで見ることができます。
ドキュメントの他の形式も、{\em docs} サブディレクトリ内で
ソースからコンパイルできます。

%%% Online copies in English and Japanese are available at
%%% \http{gnuplot.sourceforge.net/documentation.html}.
\http{gnuplot.sourceforge.net/documentation.html} に、
英語版と日本語版のコピーが置いてあります。

%%% \subsection{Worked examples}
\subsection{動作するサンプル}

%%% There is a directory of worked examples in the the source distribution.
%%% These examples and the resulting plots may also be found online at
%%% \http{gnuplot.sourceforge.net/demo/}.
ソース配布物には動作するサンプルのディレクトリが含まれています。
これらサンプル、およびその結果の描画グラフは %
\http{gnuplot.sourceforge.net/demo/} でも見ることができます。


%%% \subsection{How do I determine which options were compiled into 
%%% my \gnuplot executable?}
\subsection{私の \gnuplot にどのオプションがコンパイルされているかを
知るには}

%%% Given that you have a compiled version of \gnuplot, you can use the
%%% {\em show} command to display a list of configuration and build options
%%% that were used to build your copy.  The output formats (a.k.a. "terminals")
%%% built into your copy of gnuplot are reported by {\em set terminal}.
コンパイル済みの \gnuplot が与えられている場合、{\em show} コマンドで、
それがコンパイルされたときに使用されたコンフィグレーションと
ビルドオプションの一覧を表示できます。
あなたの gnuplot に組み込まれている出力形式 ("terminal" とも言います) は、
{\em set terminal} で表示できます。

\latex{\small}
\begin{verbatim}
gnuplot> show version long
gnuplot> set terminal
\end{verbatim}
\latex{\normalsize}


%%% \section{Working with it.}
\section{動作}

%%% \subsection{How do I get help?}
\subsection{ヘルプはどうすれば手に入れられますか}

%%% Give the {\em help} command at the initial prompt. After that, keep
%%% looking through the keywords. Good starting points are {\em help plot}
%%% and {\em help set}.
最初のプロンプトで {\em help} コマンドを実行してください。
その後キーワードをたどっていってください。
{\em help plot} や {\em help set} から始めるのがいいでしょう。

\begin{itemize}
\item
%%% Read the manual, if you have it.
もし、持っているならマニュアルを読んでください

\item
%%% Run the demos in the {\em demo} subdirectory or look at the
%%% online copies. They should give you some ideas.
%%% \http{gnuplot.info/demo}
{\em demo} サブディレクトリのデモを実行するか、
ネット上にあるそのコピーを見てください。
ヒントを与えてくれるでしょう。
\http{gnuplot.info/demo}

\item
%%% Ask your colleagues, the system administrator, or the person who
%%% set up \gnuplot.
システム管理者、または \gnuplot を設定した同僚に尋ねてください。

\item 
%%% There is an old-school usenet group devoted to gnuplot questions
%%% \news{comp.graphics.apps.gnuplot}
%%% inhabited mostly by users who found it last century.
gnuplot に関する質問を議論する Usenet グループ %
\news{comp.graphics.apps.gnuplot}
が古くからありますが、そこの住民はほとんど前世紀の人々です。

\item
%%% A more browser-oriented help forum may be found on StackOverflow
%%% \http{stackoverflow.com/questions/tagged/gnuplot}
よりブラウザ指向のヘルプフォーラムが StackOverflow にあります。
\http{stackoverflow.com/questions/tagged/gnuplot}

\item
%%% If all these fail, post a question to \news{comp.graphics.apps.gnuplot}
%%% or send mail to the gatewayed mailing list 
%%% \mailto{gnuplot-info@lists.sourceforge.net}.
%%% Please note that due to the overwhelming amount of spam it would 
%%% otherwise receive, you must subscribe before you can post to it. 
%%% Subscription instructions may be found at
%%% \http{lists.sourceforge.net/lists/listinfo/gnuplot-info}.
それらにすべて失敗した場合、質問をニュースグループ %
\news{comp.graphics.apps.gnuplot} %
に投書するか、またはメーリングリスト %
\mailto{gnuplot-info@lists.sourceforge.net} にメールを送ってください。
そのメーリングリストが圧倒的な量の spam を別に受けてしまうのを避けるため、
投稿の前にそのメーリングリストに参加 (subscribe) しなければいけない
ことに注意してください。参加の方法は、
\http{lists.sourceforge.net/lists/listinfo/gnuplot-info}
にあります。

\end{itemize}
%%% When asking a question, always mention the \gnuplot version and
%%% operating system you are using.  If you are asking about a plot that
%%% did not come out the way you expected, please try to show a minimal
%%% set gnuplot commands that produce a plot showing the problem.
質問をする際は、あなたが使用している \gnuplot のバージョンと
オペレーティングシステムについても言及する必要があります。
あなたのやり方で期待したようにはうまくできなかったグラフについて
質問したいときは、
その問題を持つグラフを生成するような最小の gnuplot コマンド列を
示すようにしてください。


%%% \subsection{How do I print out my graphs?}
\subsection{どうすればグラフをプリントアウトできますか}

%%% The output format produced by a plot command is determined by a
%%% prior {\em set terminal} command.  For non-interactive output
%%% you should pair this with a {\em set output} command to provide
%%% a file name.
plot コマンドが生成するものの出力形式は、
その前の {\em set terminal} によって決定します。
対話型でない出力に対しては、ファイル名を指定するために
これと {\em set output} コマンドを組にして実行します。

%%% As an example, the following session first plots a graph of sin(x) 
%%% to the screen and then redraws that same plot as a PDF output file.
%%% Note: the PDF plot may not look exactly like the plot on the screen.
以下の例の作業は、まず sin(x) のグラフをスクリーン上に描画し、
同じグラフを PDF 出力ファイルとして再描画します。
注意: PDF 出力は、スクリーン上に表示されるグラフと完全には
同じにならない可能性があります。

%%% \small
%%% \begin{verbatim}
%%% gnuplot> plot sin(x)
%%% gnuplot> set terminal pdf
%%% Terminal type is now 'pdfcairo'
%%% Options are ' transparent enhanced fontscale 0.5 size 5.00in, 3.00in '
%%% gnuplot> set output "sin.pdf"
%%% gnuplot> replot
%%% gnuplot> unset output       # close output file (otherwise it stays open)
%%% gnuplot> unset terminal     # return to default interactive terminal
%%% gnuplot>
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
gnuplot> plot sin(x)
gnuplot> set terminal pdf
Terminal type is now 'pdfcairo'
Options are ' transparent enhanced fontscale 0.5 size 5.00in, 3.00in '
gnuplot> set output "sin.pdf"
gnuplot> replot
gnuplot> unset output       # 出力ファイルを閉じる (でないと開いたまま)
gnuplot> unset terminal     # デフォルト対話型出力形式に戻す
gnuplot>
\end{verbatim}
\latex{\normalsize}

%%% If the start point is not the default interactive terminal you
%%% can accomplish the same thing using {\em push} and {\em pop}
出発点が、デフォルト対話型出力形式でない場合は、
同じことを {\em push} と {\em pop} で達成できます。
%% end Japanese
%%% \small
%%% \begin{verbatim}
%%% gnuplot> set terminal push    # save current terminal type (may not 
%%%   be default)
%%% gnuplot> set terminal pdf
%%% gnuplot> set out 'a.pdf'
%%% gnuplot> replot
%%% gnuplot> unset out
%%% gnuplot> set term pop         # restore saved terminal type
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
gnuplot> set terminal push   # 現在の出力形式を保存 (デフォルでなくてもよい)
gnuplot> set terminal pdf
gnuplot> set out 'a.pdf'
gnuplot> replot
gnuplot> unset out
gnuplot> set term pop        # 保存した出力形式を復帰
\end{verbatim}
\latex{\normalsize}

%%% Some interactive terminal types ({\em win, wxt, qt}) provide a 
%%% printer icon on their toolbar. This widget prints the current 
%%% plot or saves it to file using generic system tools rather than 
%%% by using a different gnuplot terminal type.
%%% That is, the file you get by selecting "save to png" 
%%% in the print menu may be different than the file you get from 
%%% {\em set term png; replot;}.
%%% In general a plot saved in this way will more closely reproduce 
%%% the screen image than a plot generated from the command line 
%%% by changing the terminal type.
対話型出力形式の中にはそのツールバー上にプリンタアイコンを提供する
ものもあります ({\em win, wxt, qt} 等)。
それは、他の出力形式を使用するのではなく、
システムに用意されている一般的なシステムツールを使い、
現在のグラフを印刷するか、またはファイルに保存します。
すなわち、印刷メニューで "save to png" を選択して得られるファイルは、
{\em set term png; replot;} によって得られるファイルとは
違うものになります。
一般に、この方法で保存したグラフは、コマンドラインから出力形式を変更して
作成したものよりも、よりスクリーン画像に近いものを生成するでしょう。


%%% \subsection{How do I include my graphs in my favorite word processor?}
\subsection{どうすればグラフを私のワープロソフトに取り込めますか}

%%% Basically, you save your plot to a file in a format your word
%%% processor can understand, and then you read in the plot from your 
%%% word processor. Vector formats (PostScript, emf, svg, pdf, \TeX{},
%%% \LaTeX{}) should be preferred,
%%% as you can scale your graph later to the right size.
基本的に、グラフをそのワープロソフトが理解できる形式のファイルに保存すれば、
そのワープロソフトで読み込むことができるでしょう。
後で正しいサイズに変換もできるという点で、ベクトル形式のフォーマット %
(PostScript, emf, svg, pdf, \TeX{}, \LaTeX{}) の方がいいでしょう。

%%% Use {\em set term} to get a list of available file formats.
使用できるファイル形式の一覧を見るには {\em set term} としてください。

%%% Many word processors can use Encapsulated PostScript (*.eps) for graphs. 
%%% You can generate eps output in \gnuplot using either
%%% {\em set terminal postscript eps}
%%% or
%%% {\em set terminal epscairo}
%%% .
%%% \Gnuplot does not embed a bitmap preview image in the output eps file.
%%% To accommodate some word processors you may have to add a preview 
%%% image yourself using an external tool before importing it into the
%%% word processor.
多くのワープロソフトは Encapsulated PostScript (*.eps) のグラフを取り込めます。
\gnuplot では、\\
{\em set terminal postscript eps} 
または
{\em set terminal epscairo}
を使うことで eps 出力を選択できます。
\gnuplot は、出力する EPS ファイルにプレビュー用のビットマップ画像を
埋め込みませんので、ワープロソフトに対応させるために、
そのファイルをワープロソフトに取り込む前に、外部ツールを使って
自分自身でプレビュー画像を追加しないといけない場合もあるかもしれません。

%%% Some applications, including the LibreOffice and Microsoft Office 
%%% suites, can handle
%%% vector images in EMF format. These can be either produced by the emf 
%%% terminal, or by selecting 'Save as EMF...' from the toolbar of the
%%% windows terminal plot window.
LibreOffice や Microsoft Office 等のアプリケーションには、
EMF 形式のベクトル画像を処理できるものがあります。
そのような画像ファイルは emf 出力形式でも作れますし、
windows 出力形式の描画ウィンドウのツールバーから %
'Save as EMF...' を選択することでも作れます。

%%% LibreOffice can also read SVG, as well as AutoCAD's dxf format.
LibreOffice は SVG 形式も、AutoCAD の dxf 形式も読みこむことができます。

%%% There are many ways to use gnuplot to produce graphs for inclusion in a
%%% \TeX\ or \LaTeX\ document.  
%%% Some terminals produce *.tex fragments for direct inclusion; others 
%%% produce *.eps, *.pdf, *.png output to be included using the 
%%% \textbackslash{}includegraphics command.
%%% The epslatex and cairolatex terminals produce both a graphics
%%% file (*.eps or *.pdf) and a *.tex document file that refers to it.
%%% The tikz terminal produces full text and graphics to a pdf file
%%% when the output is processed with pdflatex.
gnuplot を使って \TeX\ や \LaTeX\ 文書に取り込むためのグラフを作る方法は
何通りもあります。
取り込む *.tex ファイルの断片を直接生成する出力形式もありますし、
\textbackslash{}includegraphics コマンドを使って取り込むための %
*.eps, *.pdf, *.png 出力を生成するものもあります。
epslatex と cairolatex 出力形式は、画像ファイル (*.eps か *.pdf) と
それを参照する *.tex 文書ファイルの両方を生成します。
tikz 出力形式は、テキストとグラフを生成し、
その出力を pdflatex で処理すれば PDF ファイルが作成できます。

%%% Most word processors can import bitmap images (png, pbm, etc).
%%% The disadvantage of this approach is that the resolution of your
%%% plot is limited by the size of the plot at the time it is generated
%%% by \gnuplot, which is generally a much lower resolution than the
%%% document will eventually be printed in.
大抵のワープロソフトはビットマップ画像 (png, pbm 等) %
を取り込むことができますが、
この方法の短所は、
そのグラフが \gnuplot によって生成されたときのサイズによって
その解像度が制限されてしまうことです。
一般に、文書が最終的にプリントアウトされる解像度よりも
その解像度はかなり低くなります。


%%% \subsection{How do I edit or post-process a \gnuplot graph?}
\subsection{どうすれば \gnuplot のグラフを編集や後処理できますか}

%%% This depends on the terminal type you use.
これはあなたが使う出力形式に依存します。

\begin{itemize}

%%% \item \textbf{svg} terminal (scalable vector graphics) output can
%%% be further edited by a svg editor, e.g. 
%%% \textbf{Inkscape} (\http{www.inkscape.org}),
%%% \textbf{Skencil} (\http{www.skencil.org}) or
%%% \textbf{Dia} (\http{projects.gnome.org/dia/}), or loaded
%%% into \textbf{OpenOffice.org} with an on-fly conversion into OO.o Draw
%%% primitives.
\item \textbf{svg} 出力形式 (scalable vector graphics) の出力は %
\textbf{Inkscape} (\http{www.inkscape.org}),
\textbf{Skencil} (\http{www.skencil.org}), 
\textbf{Dia} (\http{projects.gnome.org/dia/}), %
のような svg エディタで更なる編集も行なえますし、
\textbf{OpenOffice.org} (OO.o Draw プリミティブへの変換によって) %
の中に取り込むこともできます。

%%% \item PostScript or PDF output can be edited directly by tools such
%%% as Adobe Illustrator or Acrobat, or can be converted to a variety
%%% of other editable vector formats by the \textbf{pstoedit} package.
%%% Pstoedit is available at \http{www.pstoedit.net}.
\item PostScript や PDF 出力は、Adobe Illustrator や Acrobat %
のようなツールで直接編集できますし、
\textbf{pstoedit} パッケージで他の色々な編集可能なベクトル形式に
変換することも可能です。
pstoedit は \http{www.pstoedit.net} にあります。

%%% \item The DXF format is the AutoCAD's format, editable by several
%%% other applications.
\item DXF 出力形式は AutoCAD の書式で、他のいくつかのアプリケーションでも
編集できます。

%%% \item Bitmapped graphics (e.g. png, jpeg, pbm) can be edited using
%%% tools such as ImageMagick or Gimp.  
\item ビットマップ形式の画像 (png, jpeg, pbm 等) は ImageMagick や %
Gimp の様なツールで編集することができます。
%% end Japanese
%%% In general, you should use a vector graphics program to post-process
%%% vector graphic formats, and a pixel-based editing program 
%%% to post-process pixel graphics.
一般に、ベクトルグラフィック形式の後処理を行なうには
ベクトルグラフィックプログラム (ドロー系ツール) を、
ピクセル画像にはピクセル画像系のプログラム (ペイント系ツール) %
を使う必要があります。

\end{itemize}

%%% \subsection{How do I save and restore my current settings?}
\subsection{どうすれば現在の設定を保存/再利用できますか}

%%% Use the {\em save "filename"} and {\em load "filename"} commands.
それには {\em save "filename"} と {\em load "filename"} のコマンドを
使ってください。

%%% \subsection{Can I put both commands and data into a single file?}
\subsection{一つのファイルに gnuplot のコマンドとデータの両方を書けますか}

%%% \Gnuplot Version 5 supports named blocks of data in "here document" 
%%% format:
\gnuplot のバージョン 5 では、「ヒアドキュメント」形式での
名前付きデータブロックをサポートしています:
%% end Japanese
\latex{\small}
\begin{verbatim}
gnuplot> $DATABLOCK << EOD
  cats 4 2
  dogs 1 4
EOD
gnuplot> plot $DATABLOCK using 2:3:1 with labels
\end{verbatim}
\latex{\normalsize}
%%% Once the named block has been defined, it can be used as many times
%%% as you like.
名前付きデータブロックは一度定義すれば、それを何度でも使うことができます。

%%% Data can also be provided in-line as part of a plot command using the
%%% pseudo-file "-".  In this case the data can only be used once.
データは、plot コマンド上で疑似ファイル指定 \verb+plot "-"+ を
行ってインライン形式で与えることもできます。
この場合は、データは一度しか使えません。

\latex{\small}
\begin{verbatim}
gnuplot> plot "-"
1 1
2 4
3 9
e
\end{verbatim}
\latex{\normalsize}

%%% \subsection{How do I run my data through a filter before plotting?}
\subsection{どうすれば描画前にデータにフィルタをかけられますか}

%%% If your operating system supports the popen() function, you
%%% can filter input data through another program or system utility
%%% as part of the {\em plot} command.
あなたの使用する OS が popen() 関数をサポートしているなら、
他のプログラムやシステムユーティリティを {\em plot} コマンドの
一部分として使用して入力データをフィルタリングすることができます。

%%% \small
%%% \begin{verbatim}
%%% gnuplot> plot "< sort +2 file.in"  # pre-sort data on column 2
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
gnuplot> plot "< sort +2 file.in"  # 2 列目のデータで事前にソート
\end{verbatim}
\latex{\normalsize}

%%% This mechanism is particularly powerful in combination with the
%%% unix-derived command line utilities {\em awk}, {\em sort} and {\em grep}.
この仕組みは、Unix 由来のコマンドラインツール {\em awk}, {\em sort}, 
{\em grep} などと組み合わせて使用するととても強力です。

%%% \subsection{Can I use \gnuplot routines for my own programs?}
\subsection{私のプログラムで \gnuplot のルーチンを利用できますか}

%%% On systems supporting pipes, you can pipe commands to \gnuplot from 
%%% other
%%% programs. Many applications with gnuplot as the graphics engine, like 
%%% Octave
%%% (\http{www.octave.org}), uses this method. This also works from a 
%%% cgi script to
%%% drive \gnuplot from a forms-based web page.
パイプをサポートしているシステムでは、
他のプログラムからパイプ経由で \gnuplot にコマンドを渡すことができます。
Octave (\http{www.octave.org}) のように、
gnuplot をグラフィックエンジンとして使用している多くのアプリケーションが
この方法を使用しています。
これは、form ベースの Web ページから \gnuplot を起動する cgi スクリプトでも
利用できます。


%%% \section{Customizing the appearance of your plot}
\section{グラフの見た目のカスタマイズ}

%%% \subsection{How to inspect or change the default colors, line, and 
%%% point properties?}
\subsection{色や線や点記号のデフォルト属性を調べたり変更するには}

%%% When you issue a {\em plot} or {\em splot} command with multiple 
%%% components,
%%% \gnuplot will by default cycle through a set of colors and linetypes.
%%% You can override this by providing specific color or linetype 
%%% properties in the
%%% plot command or you can change the default sequence.
%%% Each of the commands below accepts many additional parameters
複数の要素に対する {\em plot} や {\em splot} コマンドを実行した場合、
\gnuplot は、デフォルトでは色と線種の集合を巡回的に使用します。
これは、その plot コマンド に特定の色や線種属性を指定することで
変更できますし、デフォルトの列を変更することもできます。
以下のコマンドは、それぞれ追加パラメータをたくさん指定できます。
%% end Japanese
\begin{list}{}{}
  \item 
%%% {\em test} displays the active colors, line and point properties, etc
%%%              for the current terminal type
  {\em test} は、現在の出力形式に対する有効な色、線、点の属性などを
  表示します。
%% end Japanese
  \item 
%%%   {\em set color} or {\em set monochrome} selects a pre-defined sequence.
  {\em set color} と {\em set monochrome} は、定義済みの列を選択します。
%% end Japanese
  \item 
%%%   {\em set linetype} changes the properties of an existing linetype 
%%%     or adds a new one.
  {\em set linetype} は、存在する線種の属性を変更するか、
  新しいものを追加します。
%% end Japanese
  \item 
%%%   {\em set palette} changes the color palette used for pm3d modes 
%%%     such as heat maps.
  {\em set palette} は、温度地図のような pm3d モードで使用される
  カラーパレットを変更します。
%% end Japanese
  \item
%%%   {\em set pointsize} scales all points by an additional factor
  {\em set pointsize} は、追加の倍率として
  すべての点のサイズをスケールします。
%% end Japanese
\end{list}

%%% \subsection{Hidden line or surface removal}
\subsection{陰線処理、隠面処理}

%%% There are two relevant commands. {\em set hidden3d} affects surfaces
%%% drawn using the 3D plot style {\em splot ... with lines}. It also clips
%%% line segments created by other 3D plot styles that are occluded by those
%%% surfaces.   However it does not handle plots generated in {\em pm3d} mode.
2 つの関連するコマンドがあります。
{\em set hidden3d} は、3 次元描画スタイル {\em splot ... with lines} で
描画する曲面に影響を与えます。
それは、他の 3 次元描画スタイルで生成され、それらの曲面で隠される線分を
クリッピングします。
しかし、{\em pm3d} モードで生成されるグラフは処理しません。
%% end Japanese
%%% This includes styles {\em with pm3d}, {\em with zerror}, {\em with boxes},
%%% and miscellaneous plot elements generated while {\em set pm3d} is 
%%% in effect.
{\em with pm3d}, {\em with zerror}, {\em with boxes} の
スタイル、および {\em set pm3d} が有効な最中に描かれた
様々な描画要素についても同様です。
%% end Japanese
%%% Hidden surface removal for these plots is achieved instead by drawing 
%%% them
%%% in order of their distance from the viewer, controlled by
%%% {\em set pm3d depthorder}.
これらのグラフに対する隠面処理は、
見る側からの距離の順番で
それらを描画することによって代用的に達成されていますが、
その順番は、{\em set pm3d depthorder} で制御できます。

%%% \subsection{How do I force exact positions for the graph borders 
%%% on the page?}
\subsection{グラフの境界をページの特定の位置に丁度合わせるには}

%%% Specify the position of the top, bottom, left, and right borders in
%%% terms of their fractional position within the page:
上、下、左、右の境界位置を、ページ内の位置として比率で指定してください:

\latex{\small}
\begin{verbatim}
set lmargin at screen 0.05
set bmargin at screen 0.05
set rmargin at screen 0.95
set tmargin at screen 0.95
\end{verbatim}
\latex{\normalsize}

%%% \subsection{Arranging multiple plots next to each other on a single page}
\subsection{複数のグラフをひとつのページ内で隣同士に配置するには}

%%% The command you want is {\em set multiplot}.  The program can place
%%% a specified number of plots on a regular grid
%%% ({\em set mulplot layout <rows>, <columns> ...}) or you can position them
%%% one by one using {\em set origin} and {\em set size}.
あなたが必要とするコマンドは {\em set multiplot} です。
gnuplot は指定した数のグラフを規則的な升目状に配置します %
({\em set mulplot layout <rows>, <columns> ...})。
または、それらを {\em set origin} と {\em set size} を使って
ひとつずつ位置を決めることもできます。

%%% \subsection{How to request 1:1 scaling of axes?}
\subsection{縦と横の軸の目盛りを 1:1 にできますか}

%%% Try {\em set size square} or {\em set view equal xy}.
{\em set size square} か %
{\em set view equal xy} としてみてください。

%%% \subsection{Palette that works for both color and black\&white printing?}
\subsection{カラーとモノクロの両方の出力で機能するようなパレットは}

%%% Try {\em set palette cubehelix}.
{\em set palette cubehelix} を試してください。

%%% \subsection{How do I skip data points?}
\subsection{どうしたらデータ点をスキップできますか}

%%% By specifying \textit{?} as a data value, as in
以下のようにデータ値として \textit{?} を指定すればできます。
%% end Japanese
\latex{\small}
\begin{verbatim}
        1 2
        2 3
        3 ?
        4 5
\end{verbatim}
\latex{\normalsize}

%%% See also \textit{set missing}.
%%% See also \textit{set datafile commentschars} for specifying comment 
%%% characters in
%%% data files.
\textit{set missing} も参照してください。
データファイルのコメント文字を指定するには、
\textit{set datafile commentschars} を参照してください。


%%% \subsection{How do I plot every nth point?}
\subsection{どうしたら n 点毎にデータを扱えますか}

%%% This can be specified with various options for the command {\em plot},
%%% for example {\em plot 'a.dat' every 2}.  If you want to draw a line 
%%% through every point but only draw a point symbol at every nth point,
%%% then try {\em plot 'a.dat' with linespoints pointinterval n}.
これは、例えば {\em plot 'a.dat' every 2} のように、
{\em plot} コマンドに色々なオプションを指定することでできます。
すべての点を通る線を引き、しかし点の記号は n 点毎に書きたい場合は、\\
{\em plot 'a.dat' with linespoints pointinterval n} としてみてください。


%%% \subsection{How do I plot a vertical line?}
\subsection{どうしたら鉛直線を書けますか}

%%% Depending on context, the main methods are:
状況によりますが、主な方法は以下の通りです:
%% end Japanese
\begin{itemize}
%%% \item {\em set arrow .... .... nohead} where you have to compute
%%% explicitly the start and the end of the arrow.
%%% \item generate inlined datapoints and plot them
\item {\em set arrow .... .... nohead}  この場合、
矢の正確な開始点と終了点の位置を計算しておく必要があります。 
\item データ点を作って (インラインでの指定も可) それを描画
%% end Japanese
\end{itemize}

%%% \subsection{Y axis label in the wrong place and/or left margin too big}
\section{Y 軸のラベルの位置が悪い、あるい左の開きが大きすぎ}

%%% \Gnuplot has trouble estimating how much space to the left of the plot
%%% will be required to hold the Y axis label if it is printed horizontally
%%% ({\em set ylabel norotate}).  This is particularly true if there is
%%% TeX markup in the label string.  To work around this problem you can place
%%% the text in a numbered label rather than in {\em ylabel}.
\gnuplot は、Y 軸ラベルを水平に出力する場合 ({\em set ylabel norotate})、
それを書くのに、グラフの左にどれくらいのスペースが必要かを
評価する方法に問題を抱えていました。
これは、そのラベル文字列が TeX のマークアップを含んでいる場合は
特にそうでした。この問題の回避策として、{\em ylabel} ではなく、
番号付きラベルで文字列を置くことができます。
%% end Japanese
\latex{\small}
\begin{verbatim}
Y = 1001
set label Y '$\operatorname{\mathfrak{Im}} S_{21}$'
set label Y norotate at graph 0.0, 0.5 offset -6
\end{verbatim}
\latex{\normalsize}

%%% \section{Plot types that people often ask about}
\section{良く質問されるグラフの形式}

%%% \subsection{Animations}
\subsection{アニメーション}

%%% Only one \gnuplot terminal type (gif) directly outputs an animated file:
\gnuplot 出力形式で直接アニメーションファイルを出力できるのは
以下のもの (gif) だけです:
%% end Japanese
\begin{verbatim}
set terminal gif animate {delay <time>} {loop <N>} {optimize}
\end{verbatim}

%%% Have a look at
%%% \http{gnuplot.sourceforge.net/demo/animate2.html}
%%% in the demo collection.
デモの中の
\http{gnuplot.sourceforge.net/demo/animate2.html}
も参照してください。

%%% \subsection{Implicit defined graphs}
\subsection{陰関数のグラフ}

%%% Implicit graphs or curves cannot be plotted directly in \gnuplot.
%%% However there is a workaround. 
\gnuplot では陰関数のグラフや曲線は直接は描画できません。
しかし、次善の策はあります。
%% end Japanese
%%% \small
%%% \begin{verbatim}
%%% gnuplot> # An example. Place your definition in the following line:
%%% gnuplot> f(x,y) = y - x**2 / tan(y)
%%% gnuplot> set contour base
%%% gnuplot> set cntrparam levels discrete 0.0
%%% gnuplot> unset surface
%%% gnuplot> set table $TEMP
%%% gnuplot> splot f(x,y)
%%% gnuplot> unset table
%%% gnuplot> plot $TEMP w l
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
gnuplot> # 例: 次の行を適当な定義に置き換えてください。
gnuplot> f(x,y) = y - x**2 / tan(y)
gnuplot> set contour base
gnuplot> set cntrparam levels discrete 0.0
gnuplot> unset surface
gnuplot> set table $TEMP
gnuplot> splot f(x,y)
gnuplot> unset table
gnuplot> plot $TEMP w l
\end{verbatim}
\latex{\normalsize}
%% end Japanese
%%% The trick is to draw the single contour line z=0 of the surface
%%% z=f(x,y), and store the resulting contour curve to a temporary file or
%%% datablock.
その仕掛けは、曲面 z=f(x,y) の 1 本の等高線 z=0 を描き、
その等高線のデータを一時ファイルやデータブロックにセーブすることにあります。
 

%%% \subsection{How to fill an area between two functions}
\subsection{2 つの関数の間の領域を塗りつぶすには}

%%% A plot with filled area between two functions f(x) and g(x) can 
%%% be obtained using
%%% the pseudo file '+' with  plot style {\em filledcurves}.
2 つの関数 f(x) と g(x) の間を塗りつぶすグラフは、
疑似ファイル '+' と {\em filledcurves} を使えば簡単にできます。
%% end Japanese
\latex{\small}
\begin{verbatim}
f(x)=cos(x); g(x)=sin(x)
set xrange [0:pi]
plot '+' using 1:(f($1)):(g($1)) with filledcurves
\end{verbatim}
\latex{\normalsize}

%%% Note that this code fragment fills area between the two curves
%%% regardless of
%%% which one is above the other.  If you want to fill only the area 
%%% satisfying g(x)<f(x)
%%% or f(x)<g(x)
%%% add an additional keyword {\em above} or {\em below} after
%%%  {\em filledcurves}.
注意してもらいたいのですが、
上の例のスクリプトは 2 つの曲線の間の領域を塗りつぶすものであって、
どちらが他方の上であるかを考慮はしません。もし、
g(x)<f(x) か、または f(x)<g(x) を満たす領域のみを塗り潰したいなら、
{\em filledcurves} の後ろに {\em above} か {\em below} の
キーワードを追加してください。



%%% \subsection{Drawing a 2D projection of 3D data}
\subsection{3 次元データの 2 次元射影を描画するには}

%%% The command {\em set view map} adjusts the view angle and scaling such
%%% that subsequent 3D graphs made with {\em splot} have approximately 
%%% the same layout
%%% as a 2D graph made with {\em plot}.
コマンド {\em set view map} は、視角とスケールを調整し、
それに続く {\em splot} による 3 次元グラフを、
{\em plot} による 2 次元グラフとほぼ同じレイアウトで表示してくれます。

%%% \subsection{How to overlay dots/points scatter plot onto a pm3d map/surface}
\subsection{pm3d 色地図/曲面上に dots/points の散布図を重ねるには}

%%% Use the {\em explicit} option of the pm3d style:
pm3d スタイルの {\em explicit} オプションを使ってください:
%% end Japanese
\latex{\small}
\begin{verbatim}
gnuplot> set pm3d explicit
gnuplot> splot x with pm3d, x*y with points
\end{verbatim}
\latex{\normalsize}


%%% \subsection{How to produce labeled contours}
\subsection{ラベルつきの等高線を描画するには}

%%% Labeling individual contours in a contour plot used to require special
%%% tricks and extra processing steps in \gnuplot version 4.
%%% In version 5 it is much simpler.  Plot the contours twice,
%%% once "with lines" and once "with labels".  To make the labels stand out
%%% it may help to use
個々の等高線にラベルをつけたグラフは、\gnuplot バージョン 4 では
ある種の裏技と余分な処理手順が必要でしたが、
バージョン 5 ではその手順はとても簡単です。
等高線を、1 回目は "with lines" で、2 回目は "with labels" で %
2 回描画するだけです。
目立つラベルを作成したい場合は、以下のようにするといいでしょう。
%% end Japanese
\begin{verbatim}
set style textbox opaque noborder
set contours
splot 'DATA' with lines, 'DATA' with labels boxed
\end{verbatim}

%%% \subsection{Does \gnuplot support bar-charts/histograms/boxes?}
\subsection{\gnuplot は棒グラフ/ヒストグラム/箱型グラフが書けますか}

%%% \Gnuplot{} supports various clustered and stacked histogram styles 
%%% to display
%%% pre-tabulated data.  It also offers a few options for accumulating 
%%% raw data
%%% into bins, which can in turn be displayed as a bar chart.  See the 
%%% documentation
%%% for \textit{bins} and for \textit{smooth frequency}.
\gnuplot{} は、あらかじめ表形式にされたデータを表示するための、
さまざまな横並び型 (clustered) と積み上げ型 (stacked) の
ヒストグラムスタイルをサポートしています。
また、生のデータを場所毎に累積するためのいくつかのオプションも提供していて、
それを切り替えて棒グラフとして表示させることも可能です。
マニュアルの \textit{bins} と \textit{smooth frequency} の部分を
参照してください。

%%% \subsection{Does \gnuplot support pie charts? quarterly time charts?}
\subsection{\gnuplot は円グラフが書けますか、四半期グラフはどうですか}

%%% Pie charts are sort of difficult in \gnuplot, but see
%%% \http{gnuplot.sourceforge.net/demo/circles.html}, 
%%%  or have a look at
%%% \http{gnuplot-tricks.blogspot.com/2009/08/pie-charts-entirely-in-gnuplot.html}.
円グラフは \gnuplot ではちょっと面倒ですが、以下を見てください。
\http{gnuplot.sourceforge.net/demo/circles.html} 
または、以下を参照してください。
\http{gnuplot-tricks.blogspot.com/2009/08/pie-charts-entirely-in-gnuplot.html}

%%% The demo collection contains an example of a simple Gantt chart.
デモの中には単純なガントチャートの例があります。

%%% \subsection{Does \gnuplot support multiple y-axes on a single plot?}
\subsection{\gnuplot は一つのグラフで複数の y 軸を使えますか}

%%% Yes. 2D plots can have separate x axes at the bottom (x1) and top (x2),
%%% and separate y axes at the left (y1) and right (y2).  Version 5 offers a
%%% plot mode \textit{with parallelaxes} that allows additional y axes to be
%%% defined.
はい。2 次元グラフでは別々に x 軸を下 (x1) と上 (x2) に
持つことができますし、別々に y 軸を左 (y1) と右 (y2) に持つことができます。
バージョン 5 では、\textit{with parallelaxes} という描画モードで
y 軸をさらに追加することもできます。

%%% \subsection{How to draw a solid made of triangular facets?}
\subsection{小 3 角面で構成される物体を描画するには}

%%% It is possible to describe colored facets of a 3D object in a file
%%% organized like this:
3 次元物体の面に色をつけたものをファイルに記述するには、
以下のような形式にすれば可能です。
%% end Japanese
%%% \small
%%% \begin{verbatim}
%%% # triangle 1
%%% x0 y0 z0 <c0>
%%% x1 y1 z1 <c1>
%%% 
%%% x2 y2 z2 <c2>
%%% x2 y2 z2 <c2>
%%% 
%%% 
%%% # triangle 2
%%% x y z
%%% ...
%%% \end{verbatim}
%%% \normalsize
%% begin Japanese
\latex{\small}
\begin{verbatim}
# 三角形 1
x0 y0 z0 <c0>
x1 y1 z1 <c1>

x2 y2 z2 <c2>
x2 y2 z2 <c2>


# 三角形 2
x y z
...
\end{verbatim}
\latex{\normalsize}
%% end Japanese

%%% Notice the single and double blank lines. Also notice that the triangle
%%% really has four vertices of which two are identical.  This is because the
%%% pm3d code only knows how to deal with quadrangles.
%%% \textit{<cN>} is an optional color.
1 行の空行と 2 行の空行に注意してください。
また、三角形は実際には 4 つの頂点を持ち、そのうちの 2 つが同一である
ことにも注意してください。
これは、pm3d のコードが、四辺形を処理する方法しか知らないからです。
\textit{<cN>} はオプションの色指定です。

%%% Then plot it by (either of splot's):
これを以下のようにして描画します (どちらかの splot で):
%% end Japanese
\latex{\small}
\begin{verbatim}
set pm3d
set style data pm3d
set pm3d depthorder
splot 'facets.dat'
splot 'facets_with_color.dat' using 1:2:3:4
\end{verbatim}
\latex{\normalsize}

%%% Gnuplot is not a 3D modeling program.
%%% The depthorder rendering will not handle inter-penetrating facets.
%%% For true 3D rendering you would be probably be better off using a 
%%% ray-tracing program.
gnuplot は 3 次元モデリングソフトではありませんので、
depthorder による描画では、面同士の貫通は処理されません。
本当の 3 次元描画のためには、
多分レイトレーシングソフトを使う方がいいでしょう。

%%% \subsection{How do I plot two functions in non-overlapping regions?}
\subsection{どうしたら 2 つの関数を重ならないように書けますか}

%%% Give the desired range immediately before each function being plotted.
%%% For example, to plot experimental data and two different functional
%%% models f1 and f2 covering two different portions of the domain:
描画する各関数の直前に描かせたい範囲を置いてください。
例えば、実験データと、領域の別の部分をカバーする %
2 つの異なる関数モデル f1, f2 を描画するには:
%% end Japanese
%%% \small
%%% \begin{verbatim}
%%% gnuplot> set autoscale x  # get x range from the data
%%% gnuplot> plot 'data', [*:0] f1(x), [0:*] f2(x)
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
gnuplot> set autoscale x  # x の範囲をデータから取得
gnuplot> plot 'data', [*:0] f1(x), [0:*] f2(x)
\end{verbatim}
\latex{\normalsize}

%%% \subsection{How do I plot lines (not grids) using splot?}
\subsection{どうしたら splot で格子でない曲線を書けますか}

%%% If the data input to splot is arranged such that each line contains
%%% the same number of data points (using blank lines as delimiters, as usual),
%%% splot will by default treat the data as describing a surface.
%%% If you want to draw individual lines instead, try some combination of
%%% \textit{unset surface}, \textit{set surface explicit}, 
%%% \textit{plot ... nosurface}.
splot への入力データが、各行が同じデータ点数 (通常は空行で分離されている) %
を持つように揃っていれば、splot はデフォルトではそのデータを
曲面を記述するものとして扱います。
そうでなく個々の曲線を描画したければ、
\textit{unset surface}, \textit{set surface explicit}, 
\textit{plot ... nosurface} などを組み合わせてみてください。


%%% \subsection{How do I plot a function f(x,y) that is bounded by other
%%%           functions in the x-y plane?}
\subsection{どうしたら x-y 平面上で他の関数に制限されている f(x,y) の
          グラフを書けますか}

%%% Here is one way:
以下は一つの方法です:
%% end Japanese
\latex{\small}
\begin{verbatim}
gnuplot> f(x,y) = x**2 + y **2
gnuplot> x(u) = 3*u
gnuplot> yu(x) = x**2
gnuplot> yl(x) = -x**2
gnuplot> set parametric
gnuplot> set cont
gnuplot> splot [0:1] [0:1] u,yl(x(u))+(yu(x(u)) - yl(x(u)))*v,\
       >                   f(x(u), (yu(x(u)) - yl(x(u)))*v)
\end{verbatim}
\latex{\normalsize}


%%% \section{Text formatting and special symbols}
\section{文字列の整形と特別な記号}

%%% \subsection{Text markup using "enhanced text" mode}
\subsection{"enhanced text" モードによる文字列のマークアップ}

%%% \Gnuplot version 5 defaults to "enhanced text" mode, in which text 
%%% markup is
%%% indicated by a set of special characters embedded in the text.
\gnuplot バージョン 5 では拡張文字列モード (enhanced text) が
デフォルトで、それは特殊文字を文字列内に埋めこむことによって
文字列のマークアップを示します。

\begin{center}
\begin{tabular}{|clll|} \hline
%%% \multicolumn{4}{|c|}{Enhanced Text Control Codes} \\ \hline
%%% Control & Example & Result & Explanation \\ \hline
%%% \verb~^~ & \verb~a^x~ & $a^x$ & superscript\\
%%% \verb~_~ & \verb~a_x~ & $a_x$ & subscript\\
%%% \verb~@~ & \verb~a@^b_{cd}~ & $a^b_{cd}$ &phantom box (occupies no 
%%%    width)\\
%%% \verb~&~ & \verb~d&{space}b~ & d\verb*+     +b & inserts space of 
%%%    specified length\\
%%% \verb|~| & \verb|~a{.8-}| & $\tilde{a}$ & overprints '-' on 'a', 
%%%    raised by .8\\
%%% \verb~ ~ & \verb~ ~ & ~ ~ & times the current fontsize\\
%%% \verb| | & \verb|{/Times abc}| & {\rm abc} & print abc in font Times 
%%%    at current size\\
%%% \verb| | & \verb|{/Times*2 abc}| & \Large{\rm abc} & print abc in 
%%%    font Times at twice current size\\
%%% \verb| | & \verb|{/Times:Italic abc}| & {\it abc} & print abc in 
%%%    font Times with style italic\\
%%% \verb| | & \verb|{/Arial:Bold=20 abc}| & \Large\textsf{\textbf{abc}} 
%%%    & print abc in boldface Arial font size 20\\
%%% %% 5.3 only %% \verb|\U+| & \verb|\U+221E| & $\infty$ & Unicode 
%%% %%   point U+221E INFINITY\\
\multicolumn{4}{|c|}{拡張文字処理制御記号} \\ \hline
制御記号 & 例 & 結果 & 説明 \\ \hline
\verb~^~ & \verb~a^x~ & $a^x$ & 上付き文字\\
\verb~_~ & \verb~a_x~ & $a_x$ & 下付き文字\\
\verb~@~ & \verb~a@^b_{cd}~ & $a^b_{cd}$ &空ボックス (幅がない)\\
\verb~&~ & \verb~d&{space}b~ & d\verb*+     +b & 指定した長さのスペースを挿入\\
\verb|~| & \verb|~a{.8-}| & $\tilde{a}$ & 'a' の上に '-' を、現在のフォントサ\\
\verb~ ~ & \verb~ ~ & ~ ~ & イズの .8 倍持ち上げた位置に重ね書き\\
\verb| | & \verb|{/Times abc}| & {\rm abc} & Times フォント、今のサイズで abc を出力\\
\verb| | & \verb|{/Times*2 abc}| & \Large{\rm abc} & Times フォント、今の倍のサイズで abc\\
\verb| | & \verb|{/Times:Italic abc}| & {\it abc} & Times フォント、イタリック体で abc\\
\verb| | & \verb|{/Arial:Bold=20 abc}| & \Large\textsf{\textbf{abc}} & Arial フォント、太字、サイズ 20 で abc\\
%% 5.3 のみ %% \verb|\U+| & \verb|\U+221E| & $\infty$ & ユニコード U+221E 無限大\\
\hline
\end{tabular}
\end{center}

%%% \subsection{How do I turn text markup on or off?}
\subsection{文字列のマークアップをオンに/オフにするには}

%%% To exempt a particular text string from this processing mode use 
%%% the keyword
%%% {\em noenhanced}.  For example we don't want to interpret file names 
%%% as subscripts:
この処理モードから、特定のテキスト文字列を除外するには、
キーワード {\em noenhanced} を使用してください。
例えば、ファイル名を下付き文字と解釈させないためには:
%% end Japanese
\latex{\small}
\begin{verbatim}
set title 'Compare file_1.dat and file_2.dat' noenhanced
\end{verbatim}
\latex{\normalsize}

%%% \subsection{Is UTF-8 the answer to all my special character problems?}
\subsection{特殊文字に関するすべて問題の回答は UTF-8 にすることですか}

%%% Yes.
ええ。

%%% Unfortunately there are some circumstances where it very cumbersome 
%%% to use,
%%% notably in the generation of PostScript output.
%%% If you are working in a UTF-8 computing environment, you probably do 
%%% not have to
%%% do anything special in gnuplot to use it.
%%% If you are not, then you can still tell gnuplot to use UTF-8 when 
%%% creating output:
%%% {\em set encoding utf8}.
不運にも、特に PostScript 出力の生成時など、
使用するにはかなり厄介な環境がいくつかあります。
UTF-8 のコンピュータ環境で作業をしている場合、
gnuplot でそれを使用する場合には、多分何か特別なことを
する必要はないでしょう。
しかしそうでない場合は、出力する際には gnuplot に UTF-8 を
使用することを以下のように伝える必要があります:
{\em set encoding utf8}

%%% If you cannot enter UTF-8 characters on your
%%% keyboard you will have to solve that outside of gnuplot, or use 
%%% octal escape
%%% sequences to type them in byte-by-byte.  If your keyboard does 
%%% generate UTF-8
%%% but you don't know what keystrokes would produce a specific 
%%% character, there
%%% are probably suitable character selection apps for your desktop
%%% (e.g. KDE {\em kcharselect}, or GNOME {\em Character Map}).
UTF-8 文字列をあなたのキーボードで入力できない場合は、
gnuplot の外でそれを解決するか、
またはその 8 進のエスケープ文字列を 1 バイト毎入力してください。
あなたのキーボードで UTF-8 を生成できるが
どのキーストロークで特定の文字列を生成できるかを知らない場合は、
多分適当な文字選択アプリケーションがあなたのデスクトップに
ついているでしょう (例えば KDE なら {\em kcharselect}、
GNOME なら {\em Character Map})。

%%% \subsection{What if I need h-bar (Planck's constant)?}
\subsection{Planck 定数 (h に横線) が必要なときはどうしたらいいですか}

%%% The most straightforward way is to use a UTF-8 font, and type in the
%%% $\hbar$ character (Unicode code point \#x210F) directly. 
最も自然な方法は、UTF-8 フォントを使い、文字 $\hbar$ (Unicode コード
ポイント \#x210F) を直接入力することです。

%%% PostScript: PostScript does not handle utf8 easily so you must use an
%%% approximation based on enhanced text markup and possibly a special
%%% Symbol font:
PostScript: PostScript では utf8 の処理は面倒なので、
拡張文字列処理 (enhanced text モード) のマークアップと、
多分特別な Symbol フォントを使った擬似的なことをやる必要があります:

%%% \verb+ @{/=56 -} {/=24 h}+ or
%%% \verb+ {/=8 @{/Symbol=24 -} _{/=14 h}}+
%%% In the latter, the "-" (a long one in /Symbol) is non-spacing and 24-pt.
%%% The 14-pt "h" is offset by an 8-pt space (which is the space preceding
%%% the "\_") but smaller, since it's written as a subscript.
%%% But these don't look too much like the hbar we're used to, since the bar
%%% is horizontal instead of sloped.  
%%% Another possibility is \verb+{/=14 @^{/Symbol=10 -}{/=14 h}}+.
\verb+ @{/=56 -} {/=24 h}+ または
\verb+ {/=8 @{/Symbol=24 -} _{/=14 h}}+
後者は 24pt (ポイント) の "-" %
(/Symbol フォントの中の長いハイフン) をスペースなしで使います。
14pt の "h" は 8pt のスペース ("\_" の前にあるスペース) %
だけずらして置きますが、
それは下付き文字として書くので小さい文字になります。
しかしそれらは通常我々が使用するプランク定数の記号とは違っていて、
横線は斜線ではなく水平線になります。
もう一つの方法は \verb+{/=14 @^{/Symbol=10 -}{/=14 h}}+ とすることです。

%%% The reduced Planck's constant can be set very easily by using the
%%% AMS-LaTeX PostScript fonts which are available from
%%% \http{www.ams.org/tex/amsfonts.html} (also included in many LaTeX
%%% distributions). \Gnuplot{} (see \verb+help fontpath+) and the
%%% PostScript interpreter (usually Ghostscript) have to know where the
%%% file \verb+ msbm10.pfb+ (or \verb+ msbm10.pfa+) resides. Use
%%% \verb+ {/MSBM10 \175}+ to produce \verb+ \hslash+ which is a "h"
%%% superimposed by a sloped bar. The standard \verb+ \hbar+ (horizontal
%%% bar) has the octal code 176. Please note that h-bar exists only as an
%%% italic type.
AMS-LaTeX の PostScript フォント%
(\http{www.ams.org/tex/amsfonts.html} にあり、
そして多くの LaTeX 配布にも含まれています) を使えば、
とても簡単に小さい Planck 定数をセットすることができますが、%
\gnuplot (\verb+help fontpath+ を参照) と、
PostScript インタープリタ (通常 Ghostscript) が
ファイル \verb+ msbm10.pfb+ (または \verb+ msbm10.pfa+) が
どこにあるかを知る必要があります。%
"h" に傾斜のついた横棒を重ねた \verb+ \hslash+ を生成するには%
\verb+ {/MSBM10 \175}+ を使ってください。
通常の \verb+ \hbar+ (水平の横棒のもの) は 8 進数の 176 です。
ただし、水平の横棒はイタリック体としてしか存在しないことに注意してください。

%%% \subsection{What if I need the Solar mass symbol?}
\subsection{太陽質量記号 (丸に点) が必要なときはどうしたらいいですか}

%%% As with Planck's constant, the most straightforward way is to use a 
%%% UTF-8 font, and type in the $\odot$ character (Unicode code point 
%%% \#x2299 ; "circled dot operator") directly. 
%%% The very similar glyph at code point \#x2609 ; "sun" may be even 
%%% better, but not many fonts provide it.
Planck 定数と同様、最も自然な方法は UTF-8 フォントを使い、文字 $\odot$ %
(Unicode コードポイント \#x2299 ; "circled dot operator") %
を直接入力することです。
非常によく似たグリフがコードポイント \#x2609 にもあります。
これは "sun" で、よりいいかもしれませんが、
それを提供するフォントは多くはありません。

%%% \subsection{How to use Greek letters or other special symbols?}
\subsection{ギリシャ文字や他の特別な記号を使うには}

%%% The old-style way is to use enhanced text mode to switch to a specialize 
%%% font,
%%% e.g. the Adobe "Symbol" font, that maps the characters you want onto ascii
%%% letters 'a', 'b', etc.  This may still be necessary for PostScript output.
%%% However a much simpler way is to select UTF-8 encoding and enter the
%%% special characters just as you would any other text.
%%% This obviates the need to change fonts and gives you access to all unicode
%%% code points including CJK character sets.  To actually print or view the
%%% files produced by \gnuplot you still need appropriate fonts installed on
%%% your computer or output device.  \Gnuplot itself does not provide fonts.
古いやり方として、特別なフォント、例えば Adobe の "Symbol" フォントに
切り替えるために拡張文字処理モード (enhanced text モード) を使用する
方法があります。それは、あなたが出力したい文字に通常の ASCII 文字を
割り当てています。これは、現在でも PostScript 出力には必要でしょう。
しかし、より簡単な方法は、UTF-8 エンコーディングを選択し、
特殊文字を他の文字列と同様にそのまま入力することです。
この方法では、フォントの変更は必要なく、また CJK 文字集合を含む
すべての Unicode コードポイントへのアクセスが可能になります。
しかし \gnuplot によって作られたファイルを実際に印刷、表示するには、
あなたのコンピュータ、あるいは出力デバイスに適切なフォントが
インストールされている必要があります。
\gnuplot 自身にはフォントは含まれていません。

%%% The various \LaTeX{} terminal types (\textit{latex, epslatex, tikz, 
%%% context, cairolatex})
%%% hand off text generation to \LaTeX{}. In this case you can 
%%% use normal \LaTeX{}
%%% markup like \verb+'\alpha_{3}'+ .
いくつかの \LaTeX{} 系出力形式 (\textit{latex, epslatex, tikz, 
context, cairolatex}) は、生成した文字列を \LaTeX{} に渡します。
この場合は、\verb+'\alpha_{3}'+ のような、
通常の \LaTeX{} 風の記述が利用できます。

%%% \subsection{How do I include accented characters?}
\subsection{どうしたらアクセント付きの文字を出力できますか}

%%% If you are stuck using a non-utf8 encoding you should use
%%% 8-bit character codes together with the appropriate encoding option
%%% to obtain accented characters like \"u or \^n in your labels.
%%% You can represent the 8-bit code with an escape sequence.
%%% See the following example:
utf8 ではないエンコーディングを使い続けるのであれば、
\"u (u ウムラウト; u の上に点 2 つのアクセント) や %
\^n のような文字をラベルに含めるには、
適切なエンコードオプションを設定し、
8 ビット文字コードを使用しなければいけません。
エスケープシーケンスを使って 8 ビットコードを表現することも可能です。
例えば次のようにします。

\latex{\small}
\begin{verbatim}
gnuplot> set encoding iso_8859_1
gnuplot> set title "M\374nchner Bierverbrauch \374ber die Jahre"
gnuplot> plot "bier.dat" u 1:2
\end{verbatim}
\latex{\normalsize}

%%% Everyone else should use UTF-8, where these are normal characters.
しかし、そうでない人は UTF-8 を使うべきで、
そこではそれらは普通の文字になります。

%%% \subsection{Can I put different text sizes into my plots?}
\subsection{異なるサイズのテキストを一つのグラフで出力できますか}

%%% Most terminal types allow you to specify a starting font face and size.
%%% The "enhanced text" mode allows you to change fonts, text sizes, bold and
%%% italic styles within a plot.
たいていの出力形式では、デフォルトフォントの名前とサイズを指定できます。
「拡張文字処理」モード ("enhanced text") では、フォント、
および文字サイズやボールド、イタリックの書体を一つのグラフ内で
変更できます。



%%% \section{Miscellaneous}
\section{その他}

%%% \subsection{Does \gnuplot support a driver for graphics format XXX?}
\subsection{\gnuplot には画像形式 XXX をサポートするドライバはありますか}

%%% To see a list of the available graphic drivers for your installation of
%%% \gnuplot, type {\em set term}.
インストール済の \gnuplot が使用できるグラフィックドライバの一覧を見るには、
{\em set term} とタイプしてください。

%%% Some graphics drivers are included in the normal distribution
%%% but are not built by default. If you want to use them, you'll
%%% have to recompile from source.
通常の配布物には含まれていても、
デフォルトでは使えないようになっているグラフィックドライバもあります。
それらが使いたければ、ソースからコンパイルし直す必要があります。

%%% \subsection{I've found a bug, what do I do?}
\subsection{バグを見つけたのですがどうしたら良いですか}

%%% First, try to see whether it actually is a bug, or whether it
%%% is a feature which may be turned off by some obscure set--command.
最初に、それが本当にバグなのか、
それともある set--command で翻すことができる特徴なのかを見極めてください。

%%% Next, see whether you have an old version of \gnuplot; if you do,
%%% chances are the bug has been fixed in a newer release.
次に、あなたが使っている \gnuplot が古い版でないか確認してください。
もしそうなら、新しい版でそのバグは取り除かれている可能性があります。

%%% The development version may already contain fixes for bugs reported
%%% since the release of the current version.
%%% Before submitting a bug report, please check whether the bug in question
%%% has already been fixed.
開発版は、現在のリリース版以後に報告されたバグの修正を
既に持っている可能性があります。レポートを送る前に、
質問のバグが既に修正されていないかチェックしてみてください。

%%% If, after checking these things, you still are convinced that there is a
%%% bug please report it using the bug-tracker at
%%% \http{sourceforge.net/p/gnuplot/bugs}.
%%% Be sure to include the version of \gnuplot (including patchlevel) and 
%%% the operating system you are running it on.
%%% It helps a lot if you can provide a simple script that reproduces the
%%% error.
これらをチェックした後でなおそれがバグであると確信したなら、
\http{sourceforge.net/p/gnuplot/bugs} の bug-tracker で
報告してください。
その際、\gnuplot のバージョン (パッチレベルも含む) とそれが動いている
オペレーティングシステムも書いてください。
特にそのエラーを再現する簡単なスクリプトを提供してもらえると
大いに助かります。

%%% The tracker on sourceforge is for reporting bugs and collecting bug fixes
%%% that will appear in a subsequent release.
%%% The various online forums are a better place to ask about
%%% work arounds or actually solving \gnuplot related problems.
souceforge の tracker は、バグのレポート、および
後のリリースに反映するためのバグの修正の収集のためのものです。
\gnuplot に関する問題に対する回避策を見つける、
あるいはそれを実際に解決するには、
インターネット上に色々存在するフォーラムの方がよい場所でしょう。

%%% \subsection{Can I do heavy-duty data processing with \gnuplot? or
%%% What is beyond \gnuplot?}
\subsection{"heavy-duty" データを \gnuplot で使えますか、
または \gnuplot に代わるものはないですか}

%%% \gnuplot by itself is not very well suited for heavy numerical computation.
%%% On the other hand it can easily handle very large data sets (millions of points).
%%% If you have a specific application where the limitation is gnuplot's speed in
%%% evaluating a non-trivial function, it might be worth coding this function in
%%% C or C++ and making it a loadable plugin that gnuplot can call.
\gnuplot それ自身は、しんどい数値計算にはあまり向いていません。
しかし、かなり大きなデータ集合 (100 万データ点程度) は簡単に処理できます。
簡単ではない関数の評価のための gnuplot の速度が制限を与えてしまうような
特定のアプリケーションを使用したい場合は、
その関数を C か C++ でコーディングして作り、
それをプラグインとして gnuplot で呼び出す、という方法もあります。

%%% Two programs you might look into if you need intensive numerical computation:
さらなる数値計算を必要とする場合は、
以下の 2 つのソフトを調べてみてください:

%%% \textbf{octave} (\http{www.octave.org})
%%% is a high-level language primarily intended for numerical computations.
%%% Octave is licensed under GPL, and in principle, it is a free Matlab clone.
%%% It provides a convenient command line interface for solving linear and 
%%% nonlinear problems
%%% numerically.  By the way, octave uses \gnuplot as its plotting
%%% engine, so you get a data-processing program on top of \gnuplot.
\textbf{octave} (\http{www.octave.org}) は、
高水準言語で、主に数値計算用です。
Octave は GPL ライセンスに従い、原則的にフリーな Matlab クローンです。
線形や非線形の問題を数値的に解くのに便利な
コマンドラインインターフェースを備えています。
なお、octave は \gnuplot を描画エンジンとして使いますので、
\gnuplot の上位のデータ処理ソフトを手に入れることにもなります。

%%% \textbf{scilab} (\http{www.scilab.org}) is another open source 
%%% alternative to \textbf{matlab}.
\textbf{scilab} (\http{www.scilab.org}) は、
別の、\textbf{matlab} に代わるオープンソースソフトです。

%%% \subsection{What if I need special function that gnuplot doesn't have?}
\subsection{gnuplot が持たない特殊関数が必要な場合は}

%%% See above. You may have to use another program or write a custom plugin.
上を見てください。
他のソフトを使うか、カスタムプラグインを書くか、のいずれかが必要です。


%%% \subsection{How to use hotkeys in my interactive terminals}
\subsection{私の対話型ターミナルでホットキーを使うには}

%%% Most of the interactive terminals support both pre-defined and user-defined
%%% hotkeys to replot, toggle plot elements, change axis scaling, and so on.
%%% Hit \textit{h} in an active gnuplot plot window to get list of hotkeys. 
%%% Read \textit{help mouse} and \textit{help bind} for more information.
ほとんどの対話型出力形式では、再描画、各要素の描画のトグル、
軸のスケールの変更などを行うために、あらかじめ設定されたホットキー、
およびユーザ定義可能なホットキーの両方をサポートしています。
現在有効な gnuplot の描画ウインドウ上で \textit{h} を打つと、
ホットキーの一覧が表示されます。
さらなる情報については、\textit{help mouse} と \textit{help bind} を
参照してください。


%%% \subsection{I want to help in developing the next version of \gnuplot.
%%% What can I do?}
\subsection{\gnuplot の次期バージョンの開発について知りたいのですが
どうしたらいいですか}

%%% Join the \gnuplot beta test mailing list by sending a mail
%%% containing the line
%%% \verb+subscribe gnuplot-beta+
%%% in the body (not the subject) of the mail to
%%% \mailto{Majordomo@lists.sourceforge.net}.
\gnuplot のベータテストメーリングリストに入ってください。
メール本文 (メールタイトルではなく) に \verb+subscribe gnuplot-beta+ %
と書いてメールを %
\mailto{Majordomo@lists.sourceforge.net} に送ってください。


%%% \section{Common problems}
\section{共通の問題}


%%% \subsection{Help! None of my fonts work.}
\subsection{助けて! 私のフォントが使えない}

%%% Gnuplot does not do font handling by itself; it must necessarily leave
%%% that to the individual device support libraries. Unfortunately, this
%%% means that different terminal types need different help in finding
%%% fonts. Here are some quick hints. For more detailed information please
%%% see the gnuplot documentation for the specific terminal type you are
%%% having problems with.
gnuplot は、それ自身ではフォント処理を行いませんので、
それは必然的に個々のデバイスをサポートするライブラリに任せることになります。
残念ながらこれは、フォントの検出においては、異なる出力形式には
異なる説明が必要だということを意味しますので、
ここでは、簡単なヒントをいくつか上げておきます。
詳しい情報については、あなたが問題にしている出力形式用の %
gnuplot のドキュメントを参照してください。

\begin{description}
%%% \item [{png/jpeg/gif}] These terminal types use the libgd support 
%%% library, which
%%% searches for fonts in the directories given in the environmental variable
%%% GDFONTPATH. Once you get libgd fontpaths sorted out, you will probably
%%% want to set a default font for gnuplot.
%%% For example: \verb+setenv GNUPLOT_DEFAULT_GDFONT verdana+
\item [{png/jpeg/gif}] 
これらの出力形式は、libgd というサポートライブラリを使用し、
これは環境変数 GDFONTPATH で与えられるディレクトリ内でフォントを探します。
そのように libgd のフォントパスを設定すると、
今度は gnuplot のデフォルトフォントを設定したいかもしれません。
例: \\
\verb+setenv GNUPLOT_DEFAULT_GDFONT verdana+
%% end Japanese
%%% \item [{post}] PostScript font names are not resolved until the document
%%% is printed. Gnuplot does not know what fonts are available to your
%%% printer, so it will accept any font name you give it. However, it
%%% is possible to bundle a font with the gnuplot output; please see the
%%% instructions given by gnuplot's internal command {}``help set term
%%% post fontfile''.
\item [{post}]
PostScript フォント名は、その文書が印刷されるまで参照されませんから、
あなたのプリンタでどのフォントが有効であるかを gnuplot は知りません。
よって、どんなフォント名でも指定することができます。
しかし、gnuplot の出力にフォントを埋め込むことは可能です。それについては、
gnuplot の内部コマンド ``help set term post fontfile'' で得られる
指示を参照してください。
%% end Japanese
%%% \item [{svg}] Font handling is viewer-dependent. 
\item [{svg}] フォント処理は、ビューワ依存です。
%% end Japanese
%%% \item [{x11}] The x11 terminal uses the normal x11 font server mechanism.
%%% The only tricky bit is that in order to use multi-byte fonts you must
%%% explicitly say so:
\item [{x11}]
x11 出力形式は、通常の x11 フォントサーバ機構を使用しますが、
少しだけトリッキーな点があり、マルチバイトフォントを使用するためには、
明示的に以下のように指定する必要があります:
%% end Japanese
\latex{\small}
\begin{verbatim}
set term x11 font "mbfont:sazanami mincho,vera,20"
\end{verbatim}
\latex{\normalsize}
%%% \item [{win}] Select "Choose font..." from the "Options" pull-down menu 
%%% in the toolbar.
\item [{win}] 
ツールバーの "Options" のプルダウンメニューから "Choose font..." を選択します。
%% end Japanese
%%% \item [{wxt, qt}] On linux systems these terminals rely on fonts provided
%%% by the system's \textit{fontconfig} utility.
\item [{wxt, qt}] 
Linux システムでは、これらの出力形式はシステムの fontconfig ユーティリティが
提供するフォントを使用します。
%% end Japanese
\end{description}

%%% \subsection{The first plot in a qt terminal session fails or has bad layout}
\subsection{qt 出力形式のセッションでの最初のグラフが失敗したり
変なレイアウトになります}

%%% You may also get an error message about "slow font initialization".
%%% This is because qt relies on a shared system font cache.
%%% If you request a font that no one else is using, it takes a while to
%%% update the cache.  This mostly happens on Windows or OSX, because other
%%% programs on those systems tend to use a different font mechanism so the 
%%% relevant font cache may be empty.
%%% Try invoking gnuplot with command line option {\em -\@-slow}.
"slow font initialization" というエラーメッセージも出たかもしれません。
これは、qt が共有システムフォントキャッシュに依存しているからです。
もし、他の誰も使っていないフォントを要求すると、
そのキャッシュの更新に時間がかかります。
これは、Windows や OSX ではよく起こるのですが、
それらのシステムでは他のソフトが異なるフォントの仕組みを
使用する傾向にあるため、関連するフォントキャッシュが空になるからです。
gnuplot の呼び出し時に、コマンドラインオプション {\em -\@-slow} を
つけてみてください。

%%% \subsection{Pm3d splot from a datafile does not draw anything}
\subsection{pm3d によるデータファイルの splot で何も表示されません}

%%% You do {\em set pm3d; splot 'a.dat'} and no plot but colorbox appears.
%%% Perhaps there is no blank line in between two subsequent 
%%% scans (isolines) in
%%% the data file? Add blank lines! If you are curious what this means, 
%%% then don't
%%% hesitate to look to files like {\em demo/glass.dat} or 
%%% {\em demo/triangle.dat}
%%% in the gnuplot demo directory.
{\em set pm3d; splot 'a.dat'} とやっても、
カラーボックスは表示されるもののグラフは表示されないとしたら、
もしかしたら 2 つのつながっている線 (孤立線) を分離する空行が
ないのでは ? 空行を入れなさい ! %
これが何を意味するのかに興味があるなら、
gnuplot の demo ディレクトリにある {\em demo/glass.dat} や %
{\em demo/triangle.dat} などのファイルをじっくり見てください。

%%% You can find useful the following awk script (call it e.g. 
%%% {\em addblanks.awk})
%%% which adds blank lines to a data file whenever number in the first column
%%% changes:
以下の awk スクリプト %
(例えば {\em addblanks.awk} と呼ぶことにします) %
は有用で、データファイルに、
第 1 列目の数が変更したときに空行を追加してくれます。
%% end Japanese
%%% \small
%%% \begin{verbatim}
%%% /^[[:blank:]]*#/ {next} # ignore comments (lines starting with #)
%%% NF < 3 {next} # ignore lines which don't have at least 3 columns
%%% $1 != prev {printf "\n"; prev=$1} # print blank line
%%% {print} # print the line
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
/^[[:blank:]]*#/ {next} # コメント行 (# で始まる行) は無視
NF < 3 {next} # 3 列未満しかない行も無視
$1 != prev {printf "\n"; prev=$1} # 空行を出力
{print} # その行自体を出力
\end{verbatim}
\latex{\normalsize}

%%% Then, either preprocess your data file by command
%%% {\em awk -f addblanks.awk <a.dat}
%%% or plot the datafile under a unixish platform by\\
%%% {\em gnuplot> splot "<awk -f addblanks.awk a.dat"}.
これにより、データファイルをコマンド
{\em awk -f addblanks.awk <a.dat}
によって前処理するか、
unix 互換のプラットフォーム上でデータファイルを \\
{\em gnuplot> splot "<awk -f addblanks.awk a.dat"}
のように描画するか、のどちらかを行ってください。

%%% \subsection{Why does \gnuplot ignore my very small numbers?}
\subsection{なぜ \gnuplot はとても小さい数字のデータを無視するのですか}

%%% For some purposes \Gnuplot{} treats numbers less than 1e-08 as being zero.
%%% Thus, if you are trying to plot a collection of very small
%%% numbers, they may be plotted as zero. Worse, if you're plotting
%%% on a log scale, they will be off scale. Or, if the whole set of
%%% numbers is "zero", your range may be considered empty:
\gnuplot は、ある目的のため 1e-08 より小さい数を 0 として扱います。
よってとても小さい数の集まりを描画しようとすると
それらは 0 として描画されてしまいます。
対数軸で描画する場合は、
もっとひどいことにそれらは目盛りからはずれてしまいます。
また、全ての数が "0" である場合、
範囲は空であるとみなされます:
%% end Japanese
%%% shige: 以前は「(訳注: \gnuplot 3.6 以前の仕様)」をつけていた

\latex{\small}
\begin{verbatim}
gnuplot> plot 'test1'
Warning: empty y range [4.047e-19:3e-11], adjusting to [-1:1]
gnuplot> set yrange [4e-19:3e-11]
gnuplot> plot 'test1'
              ^
y range is less than `zero`
\end{verbatim}
\latex{\normalsize}

%%% The solution is to change \gnuplot's idea of "zero":
その対策は \gnuplot の "zero" の概念を変更することです。
%% end Japanese
\latex{\small}
\begin{verbatim}
gnuplot> set zero 1e-20
\end{verbatim}
\latex{\normalsize}

%%% For more information, type {\em help set zero}.
より詳しい情報については {\em help set zero} を参照してください。

%%% \subsection{When I {\em replot} a multiplot I get only a fragment}
\subsection{multiplot のグラフを {\em replot} すると一かけらしか出ません}

%%% \Gnuplot only retains enough information to regenerate the most recent
%%% {\em plot} or {\em splot} command.
%%% You need to save the complete sequence of commands that generated the
%%% multiplot into a script file. Then you can {\em load} that script into
%%% \gnuplot as many times as you need for replotting the drawing to different
%%% terminals or output files.
\gnuplot は、直近の {\em plot} や {\em splot} コマンドを
再生成するために十分な情報を保持するだけなので、multiplot に対しては、
それを生成する完全なコマンド列を
スクリプトファイルに保存する必要があります。
そして、{\em load} することでその内容を \gnuplot に取り込めば、
あなたが必要とする回数だけ異なる出力形式、異なる出力ファイルに対して
再描画をすることができます。

%%% \subsection{My formulas (like 1/3) are giving me nonsense results! 
%%% What's going on?}
\subsection{書いた式 (1/3 など) が意味のない結果を出力してしまいます}

%%% \Gnuplot{} does integer, and not floating point, arithmetic on
%%% integer expressions. For example, the expression 1/3 evaluates
%%% to zero. If you want floating point expressions, supply
%%% trailing dots for your floating point numbers. Example:
\gnuplot は整数の式の場合、実数演算ではなく整数演算を行ないます。
例えば式 1/3 は 0 と評価されます。
もし実数の値が欲しいならば、その数の後にドット "." をつけてください。
例:


\latex{\small}
\begin{verbatim}
gnuplot> print 1/3
                0
gnuplot> print 1./3.
                0.333333
\end{verbatim}
\latex{\normalsize}

%%% This way of evaluating integer expressions is shared by both C and Fortran.
整数の式をこのようにして評価するやり方は %
C や Fortran でも行なわれています。


%%% \subsection{My output files are incomplete!}
\subsection{私の出力が不完全です}

%%% You may need to flush the output and close the file with
%%% {\em set output} or {\em unset output}).
{\em set output} か {\em unset output} によって、
たまっている出力を吐き出させ、そして出力をクローズする必要が
あるかも知れません。

%%% Some output formats (postscript, pdf, latex, svg, ...) can include several
%%% pages of plots in a single output file.  For these output modes, gnuplot
%%% leaves the file open after each plot so that you can add additional plots
%%% to it.  The file is not completed and made available to external applications
%%% or select a different terminal type ({\em set term}) or exit gnuplot.
出力形式によっては、一つの出力ファイルに複数のページが含まれています %
(postscript, pdf, latex, svg, ...)。
これらの出力モードでは、gnuplot はあなたが描画の追加ができるように
各描画後にファイルをオープンしたままにしています。
よって異なる出力形式を選択する ({\em set term}) か、
または gnuplot を終了するまでは、そのファイルは完全ではなく、
外部のアプリケーションでは使用できるようにはなっていません。


%%% \subsection{Calling gnuplot in a pipe or from a script doesn't 
%%% show a plot on my screen!}
\subsection{パイプやスクリプトから gnuplot を呼び出すと
スクリーンにグラフがでません}

%%% One common cause is that \gnuplot exits immediately after drawing,
%%% the plot window closes when gnuplot exits, and it all happens too fast
%%% for you see the plot.
%%% There are several solutions to this:
よくある原因の一つが、\gnuplot がグラフの描画直後に終了してしまうことで、
終了時に描画ウィンドウが閉じてしまうため、
あなたがグラフを見れないぐらいにすべての表示があっという間に
終わってしまう、という状態です。
これには、いくつかの解決法があります。
%% end Japanese
\begin{itemize}
%%% \item  Use the gnuplot command line option {\em -persist}.
%%%        This leaves the plot window open after gnuplot itself exits.
\item  gnuplot のコマンドラインオプション {\em -persist} を使用すること。
       これは、gnuplot 自身の終了後も描画ウィンドウを残します。
%% end Japanese
%%% \item  Interactive terminals also accept {\em persist} as an option
%%% to {\em set term}.
\item  オプション {\em set term} で {\em persist} を受け付ける
       対話型出力形式を使用すること。
%% end Japanese
%%% \item  Send a {\em pause mouse close} command to gnuplot before closing 
%%%        the pipe.
%%%        This will tell gnuplot not to exit until the plot window is 
%%%        closed by
%%%        some other action.
\item  パイプの終了前に gnuplot にコマンド {\em pause mouse close} を
       送信すること。これは、描画ウィンドウが他の動作で終了するまで
       gnuplot を終了させないように指示します。
%% end Japanese
%%% \item  Have the script close the pipe only when it is sure you don't need
%%%        gnuplot and its plot window any more.
\item  あなたが gnuplot、および描画ウィンドウ等々を本当に必要と
       しなくなった場合にのみ、スクリプトにパイプを閉じさせること。
%% end Japanese
\end{itemize}

%%% Here is a short Perl-script showing two of these fixes:
以下は、これらの修正のうち 2 つを示す Perl スクリプトです:
%% end Japanese
%%% \small
%%% \begin{verbatim}
%%% #!/usr/local/bin/perl -w
%%% open (GP, "|/usr/local/bin/gnuplot -persist") or die "no gnuplot";
%%% # force buffer to flush after each write
%%% use FileHandle;
%%% GP->autoflush(1);
%%% print GP,"set term x11; plot sin(x) with lines\n";
%%% print GP,"pause mouse close\n";
%%% close GP
%%% \end{verbatim}
%%% \normalsize
\latex{\small}
\begin{verbatim}
#!/usr/local/bin/perl -w
open (GP, "|/usr/local/bin/gnuplot -persist") or die "no gnuplot";
# 各出力後に強制的にバッファを掃き出す
use FileHandle;
GP->autoflush(1);
print GP,"set term x11; plot sin(x) with lines\n";
print GP,"pause mouse close\n";
close GP
\end{verbatim}
\latex{\normalsize}


%%% \section{Credits}
\section{帰属}

%%% This FAQ was initially compiled by John Fletcher with contributions
%%% from Russell Lang, John Campbell, David Kotz, Rob Cunningham, Daniel
%%% Lewart and Alex Woo. Reworked by Thomas Koenig from a draft
%%% by Alex Woo, with corrections and additions from Alex Woo, John
%%% Campbell, Russell Lang, David Kotz and many corrections from Daniel Lewart.
%% end Japanese
この FAQ は、
Russell Lang, John Campbell, David Kotz, Rob Cunningham, Daniel Lewart,
Alex Woo の寄与を John Fletcher が最初にまとめたものです。
そして、Alex Woo の草案により、
Alex Woo, John Campbell, Russell Lang, David Kotz らの修正と追加、
Daniel Lewart による多くの修正が Thomas Koenig によりまとめ直されました。
%%% Again reworked for \gnuplot 3.7 by Alexander Mai and Juergen v.Hagen
%%% with corrections by Lars Hecking, Hans-Bernhard Broecker and others.
そして、\gnuplot 3.7 用に再び %
Lars Hecking, Hans-Bernhard Broeker およびその他の人々による修正が %
Alexander Mai と Juergen v.Hagen によりまとめ直されました。
%% end Japanese
%%% shige: Broecker -> Broeker (2019-05-06)
%%% Revised for \gnuplot version 4 by Petr Mikul\'{\i}k and Ethan Merritt.
\gnuplot バージョン 4 用の修正は、
Petr Mikul\'{\i}k と Ethan Merritt により行なわれました。
%% end Japanese
%%% Revised for \gnuplot version 5 by Ethan Merritt.
\gnuplot バージョン 5 用の修正は、
Ethan Merritt により行なわれました。

%%% 
(訳注: 日本語訳は Shigeharu TAKENO (\mailto{shige@iee.niit.ac.jp}) により
行なわれました。)
%% end Japanese
\end{document}
