
%% Time-stamp: <99/01/02 23:44:25 bruce>

\documentclass[twocolumn]{article}
\usepackage{fancybox,dcolumn,amsmath,amssymb}

\setlength{\parindent}{0truecm}
\setlength{\parskip}{1ex}
\setlength{\hoffset}{-0.5truecm}
\setlength{\voffset}{0truecm}
\setlength{\topmargin}{-2.5truecm}
\setlength{\marginparsep}{0truecm}
\setlength{\marginparwidth}{0truecm}
\setlength{\textheight}{25.5truecm}
\setlength{\textwidth}{17truecm}
\setlength{\oddsidemargin}{0truecm}
\setlength{\evensidemargin}{1.2truecm}
\setlength{\columnsep}{1.4truecm}

\newenvironment{Boxedminipage}%
{\begin{Sbox}\begin{minipage}}%
  {\end{minipage}\end{Sbox}\Ovalbox{\TheSbox}}

\def\version{{0.5}}
\def\revised{{January 2, 1999}}
\def\file#1{{\texttt{`#1'}}}
\def\key#1{{\textrm \leavevmode\hbox{%
  \raise0.4pt\hbox{$\langle$}\kern-.08em\vtop{%
    \vbox{\hrule\kern-0.4pt
     \hbox{\raise0.4pt\hbox{\vphantom{$\langle$}}#1}}%
    \kern-0.4pt\hrule}%
  \kern-.06em\raise0.4pt\hbox{$\rangle$}}}}
\def\variable#1#2#3{{
    \vspace{-0.2truecm}
    \begin{flushright}
      \begin{minipage}[h]{0.97\linewidth}
        \vspace{-0.2truecm}
        \textbf{#1}\hfill[\texttt{#2}]
        \begin{flushright}
          \begin{minipage}[h]{0.93\linewidth}
            \vspace{-0.2truecm}
            #3
          \end{minipage}
        \end{flushright}
      \end{minipage}
    \end{flushright}
    }}
\def\command#1#2{{
    \vspace{-0.2truecm}
    \begin{flushright}
      \begin{minipage}[h]{0.97\linewidth}
        \vspace{-0.2truecm}
        \textbf{#1}\hfill
        \begin{flushright}
          \begin{minipage}[h]{0.93\linewidth}
            \vspace{-0.4truecm}
            #2
          \end{minipage}
        \end{flushright}
      \end{minipage}
    \end{flushright}
    }}
\def\Star{{$\star$}}


\begin{document}

\small

\thispagestyle{empty}

\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      \vspace{0.01\textheight}
      {\Large Quick Reference for}\\
      \vspace{0.007\textheight}
      {\Large gnuplot mode}
      \vspace{0.01\textheight}
    \end{center}
  \end{Boxedminipage}
\end{center}

\vspace{3ex}

This card describes the features of gnuplot mode for emacs.  Gnuplot
mode is intended for composing scripts for the \textsc{gnuplot}
plotting program.  It offers functions for sending commands or entire
scripts to the \textsc{gnuplot} program as well as various functions
to aid in composing scripts.  See the comments in the file
\file{gnuplot.el} for instructions on installing gnuplot mode.

\vspace{2ex}

\begin{minipage}[h]{\linewidth}
  \begin{center}
    \centerline{{\large\textbf{gnuplot mode key sequences}}}
    \vspace{0.2ex}
    \begin{tabular}[h]{>{\ttfamily}cl}
      \hline \hline
      \textrm{key} & \quad description \\
      \hline
      C-c C-l & send line to gnuplot \\
      C-c C-v & send line and move forward 1 line\\
      C-c C-r & send region to gnuplot \\
      C-c C-b & send buffer to gnuplot \\
      C-c C-f & send a file to gnuplot \\
      & \\[-1.5ex]
      M-\key{tab} & complete keyword at point \\
      C-c C-i & insert filename at point \\
      C-c C-n & negate set option at point \\
      C-c C-c & set arguments of command at point \\
      S-mouse-2 & set arguments of command under mouse \\
      C-c C-h & get help from the gnuplot info file \\
      C-c C-e & look at the gnuplot process buffer \\
      & \\[-1.5ex]
      C-c C-k & kill the gnuplot process \\
      C-c C-u & submit a bug report about gnuplot mode \\
      C-c C-z & customize gnuplot mode \\
      \hline \hline
    \end{tabular}
  \end{center}
\end{minipage}

\vspace{4ex}

\centerline{{\large\textbf{Starting gnuplot-mode}}}
\vspace{2ex}

\command{M-x gnuplot-mode}{Start gnuplot-mode in the current buffer.}
%
\command{M-x gnuplot-make-buffer}{Open a new buffer in gnuplot-mode}



\vfill

\begin{Boxedminipage}{0.98\linewidth}
  \begin{center}
    \footnotesize{
      Gnuplot-mode homepage \\
      \texttt{http://feff.phys.washington.edu/\char126ravel/gnuplot/}}
  \end{center}
\end{Boxedminipage}
\begin{flushleft}
  {\footnotesize
    This page {\copyright} 1998 Bruce Ravel \hfill revised \revised \\
    \texttt{<ravel@phys.washington.edu>} \hfill printed \today \\
    This page covers version {\version} of gnuplot-mode.

    Permission is granted to make and distribute copies of this quick
    reference provided the copyright notice and this permission are
    preserved on all copies.}
\end{flushleft}
%%\vfil
\pagebreak

%%% end of first column

\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      {\large User configurable variables}
    \end{center}
  \end{Boxedminipage}
\end{center}


\variable{gnuplot-program}{gnuplot}{The name of the gnuplot
  executable.}
%
\variable{gnuplot-process-name}{*gnuplot*}{The name of the gnuplot
  process and process buffer.}
%
\variable{gnuplot-gnuplot-buffer}{plot.gp}{The name of the gnuplot
  scratch buffer opened by \texttt{gnuplot-make-buffer}.}
%
%
\variable{gnuplot-display-process}{'window}{Determines how to display
  the gnuplot process buffer, either 'frame, 'window, or nil}
%
\variable{gnuplot-info-display}{'window}{Determines how
  `gnuplot-get-help' displays the info file, either 'frame, 'window,
  or nil}
%
\variable{gnuplot-which-highlight}{'font-lock}{Determines which syntax
  highlighting scheme to use, font-lock or hilit19.}
%
\variable{gnuplot-echo-command-line-flag}{nil}{If lines that you send to
  gnuplot from the gnuplot mode buffer are not appearing at the
  gnuplot prompt in the process buffer, set this to t.}
%
\variable{gnuplot-delay}{0.01}{Time in seconds to allow the gnuplot
  display to update.  Increase this number if the prompts and lines
  are displayed out of order.}
%
\variable{gnuplot-quote-character}{'}{Quotation character used when
  inserting a filename into the script (single, double, or no quote).}


%%\vspace{1ex}
\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      {\large Hook variables}
    \end{center}
  \end{Boxedminipage}
\end{center}

\variable{gnuplot-mode-hook}{nil}{Functions run when gnuplot minor
  mode is entered.}
%
\variable{gnuplot-load-hook}{nil}{Functions run when gnuplot.el is
  first loaded.}
%
\variable{gnuplot-after-plot-hook}{nil}{Functions run after gnuplot
  plots an entire buffer.  See the doc string for
  \texttt{gnuplot-recently-sent}.}
%
\variable{gnuplot-comint-setup-hook}{nil}{Functions run after setting
  up the gnuplot process buffer in comint mode.}
%
\variable{gnuplot-info-hook}{nil}{Functions run before setting up
  info-look in the gnuplot-mode buffer.}
%



\vfill\eject


\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      {\large Insertion variables}
    \end{center}
  \end{Boxedminipage}
\end{center}

\noindent These variables control the \texttt{Insertions} pull-down
menu, which can be used to insert \textsc{gnuplot} commands into the
script.  The various sub-menu variables can be used to customize which
commands appear in the \texttt{Insertions} menu.

\vspace{2ex}
%
\variable{gnuplot-insertions-menu-flag}{t}{Non-nil means to display
  the \texttt{Insertions} menu in the menubar.}
%
\variable{gnuplot-insertions-show-help-flag}{nil}{Non-nil means to
  display help from info file when using the \texttt{Insertions}
  menu.}
%
\variable{gnuplot-insertions-adornments}{\Star}{Contents of the
  \texttt{adornments} sub-menu.}
%
\variable{gnuplot-insertions-plot-options}{\Star}{Contents of the
  \texttt{plot-options} sub-menu.}
%
\variable{gnuplot-insertions-terminal}{\Star}{Contents of the
  \texttt{terminal} sub-menu.}
%
\variable{gnuplot-insertions-x-axis}{\Star}{Contents of the \texttt{x
    axis} sub-menu.}
%
\variable{gnuplot-insertions-x2-axis}{\Star}{Contents of the \texttt{x2
    axis} sub-menu.}
%
\variable{gnuplot-insertions-y-axis}{\Star}{Contents of the \texttt{y
    axis} sub-menu.}
%
\variable{gnuplot-insertions-y2-axis}{\Star}{Contents of the \texttt{y2
    axis} sub-menu.}
%
\variable{gnuplot-insertions-z-axis}{\Star}{Contents of the \texttt{z
    axis} sub-menu.}
%
\variable{gnuplot-insertions-parametric-plots}{\Star}{Contents of the
  \texttt{parametric plots} sub-menu.}
%
\variable{gnuplot-insertions-polar-plots}{\Star}{Contents of the
  \texttt{polar plots} sub-menu.}
%
\variable{gnuplot-insertions-surface-plots}{\Star}{Contents of the
  \texttt{surface plots} sub-menu.}



\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      {\large Toolbar variables}
    \end{center}
  \end{Boxedminipage}
\end{center}

\noindent These variables control the use and location of the
toolbar in XEmacs.  The toolbar has buttons equivalent to the key
sequences \texttt{C-c C-l}, \texttt{C-c C-r}, \texttt{C-c C-b},
\texttt{C-c C-e}, and \texttt{C-c C-h}.

\vspace{2ex}

%
\variable{gnuplot-display-toolbar-flag}{nil}{Non-nil means to display
  a toolbar if using XEmacs.}
%
\variable{gnuplot-use-toolbar}{left-toolbar}{Location of XEmacs
  toolbar.  Valid values are \texttt{left-toolbar},
  \texttt{right-toolbar}, \texttt{top-toolbar}, \texttt{bottom-toolbar},
  \texttt{default-toolbar} and nil.}
%

\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      {\large Set Arguments}
    \end{center}
  \end{Boxedminipage}
\end{center}

\noindent These variables control the behavior of the graphical
interface to setting command arguments.  \texttt{C-c C-c} with point
over a command or \texttt{S-mouse-2} with the mouse cursor over a
command will cause a small frame to pop open with which you can set
command arguments.  Green button with bold text are bound to pup-up
menus --- use the mouse-2 to select an item from the menu.  Grey fields
are for filling in strings or numbers.  Hit the \textbf{[Set Options]}
button with \texttt{mouse-2} to insert command arguments into the
script.  You can also use the \key{tab} key to move among the widgets
and \key{ret} to push the buttons.

\vspace{2ex}
%
\variable{gnuplot-gui-popup-flag}{nil}{When non-nil an argument
setting frame will pop open whenever the \texttt{Insertions} menu is
used.}
%
\variable{gnuplot-gui-frame-plist}{\Star}{Property list of parameters
  controlling the argument setting frame.  Used by XEmacs.}
%
\variable{gnuplot-gui-frame-parameters}{\Star}{List of parameters
  controlling the argument setting frame.  Used by Emacs.}
%
\variable{gnuplot-gui-fontname-list}{\Star}{List of font available on your
  computer to the terminal drivers.}
%


\begin{center}
  \begin{Boxedminipage}{0.75\linewidth}
    \begin{center}
      {\large Faces}
    \end{center}
  \end{Boxedminipage}
\end{center}

\noindent These are various faces defined for use with gnuplot mode.
\vspace{2ex}

%
\variable{gnuplot-prompt-face}{firebrick}{Color of gnuplot prompt (on a
  light background) in process buffer.  Bold and underlined on a
  monochrome display.}
%
\variable{gnuplot-gui-menu-face}{dark olive green}{Color of menu
  buttons (on a light background) in the argument setting frame.
  Italic on a monochrome display.}
%
\variable{gnuplot-gui-button-face}{sienna}{Color of push buttons (on a
  light background) in the argument setting frame.  Italic on a
  monochrome display.}
%
\variable{gnuplot-gui-label-face}{dark slate blue}{Color of buttons (on
  a light background) used to set label lists in the argument setting
  frame.  Italic on a monochrome display.}
%

\vfill
\hrule
\vspace{0.5ex}
\begin{flushleft}
  \footnotesize{Variables marked with {\Star} have default values that
    are too long to print here.}
\end{flushleft}

\end{document}




%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
